\section{Introduction}
\label{sec:intro}

Previous work by Dunfield~\cite{dunfield2014elaborating} has shown the
usefulness of type systems with intersection types and a merge operator. The presence of a merge
operator in a core calculus provides significant expressiveness, allowing
encodings for many other language constructs as syntactic sugar. For example
single-field records are easily encoded as types with a label, and multi-field
records are encoded as the concatenation of single-field records. Concatenation
of records is expressed using intersection types at the type-level and the
corresponding merge operator at the term level. Dunfield formalized a simply
typed lambda calculus with intersection types and a merge operator. He showed
how to give a semantics to the calculus by a type-directed translation to a
simply typed lambda calculus extended with pairs. The type-directed translation
is elegant and type-safe.

% \bruno{the following paragraph no longer makes sense for this paper.}
% Intersection types and the merge operator are also useful in the
% context of software \emph{extensibility}. In recent years there has
% been a wide interest in presenting solutions to the \emph{expression
%   problem}~\cite{wadler1998expression} in various
% communities. Currently there are various solutions in functional
% programming languages~\cite{swierstra:la-carte,emgm}, object-oriented
% programming
% languages~\cite{togersen:2004,Zenger-Odersky2005,oliveira09modular,oliveira2012extensibility}
% and theorem provers~\cite{DelawareOS13,SchwaabS13}.
% Many of the proposed solutions for extensibility are closely related
% to type-theoretic encodings of datatypes~\cite{BoehmBerarducci},
% except that some form of subtyping is also involved.  Various
% language-specific mechanisms are used to combine ideas from
% type-theoretic encodings of datatypes with subtyping, but the essence
% of the solutions is hidden behind the peculiarities of particular
% programming languages.  Calculi with intersection types have a natural
% subtyping relation that is helpful to model problems related to
% extensibility.  Moreover, intersection types and an \emph{encoding} of
% a merge operator have been shown to be useful to solve additional
% challenges related to extensibility~\cite{oliveira2013feature}.
% Therefore it is natural to wonder if a
% core calculus supporting parametric polymorphism,
% intersection types and a merge operator, can be used to capture the
% essence of various solutions to extensibility problems.

%Clearly this seems to indicate that a more
%foundational approach is lacking. What is currently missing is a
%foundational core calculus that can capture the key ideas behind the
%various solutions.

While Dunfield's calculus is type-safe, it lacks the property of
\emph{coherence}: different derivations for the same expression can
lead to different results. The lack of coherence is an important
disadvantage for adoption of his core calculus in implementations of
programming languages, as the semantics of the programming language
becomes implementation dependent. Although Dunfield mentioned the
possibility of extending the type system to allow only disjoint
intersection types, he did not formalize or further pursue this
approach.


\begin{comment}
Moreover, from the theoretic point-of-view, the ambiguity that arises
from the lack of coherence makes the calculus unsatisfying when the
goal is to precisely capture the essence of solutions to
extensibility.
\end{comment}

This paper presents \name: a core calculus with a variant of \emph{intersection
types} and a \emph{merge operator}. The semantics of \name is both type-safe and
coherent. Coherence is achieved by ensuring that intersection types are
\emph{disjoint}. \bruno{introduction needs to be rewritten at least from this point!}

Given two types $A$ and $B$, two types are disjoint ($A *
B$)\footnote{The notation $A * B$ is inspired by the \emph{separating
conjunction} construct in Reynolds' separation
logic~\cite{reynolds2002separation}.} if there is no type $C$ such that both $A$
and $B$ are subtypes of $C$. Formally this definition is captured as follows:
\[A * B \equiv \not\exists C.~A <: C \wedge B <: C\] With this definition of
disjointness we present a formal specification of a type system that prevents
intersection types that are not disjoint. However, the formal definition of
disjointness does not lend itself directly to an algorithmic implementation.
Therefore, we also present an algorithmic specification to determine whether two
types are disjoint. Moreover, this specification is shown to be sound and
complete with respect to the formal definition of disjointness. We have
mechanized the many of the interesting metatheoretical results using the Coq proof
assistant, and the main proof of type-safety and the coherence theorem can be found
in the full version of the paper\footnote{\url{https://github.com/zhiyuanshi/intersection/blob/master/esop-16/paper-full.pdf}}.

As an application of \name, we show that \name can serve as a
foundation for a trait-based~\cite{scharli2003traits} object-oriented
language. Due to the type system of \name, the language supports
common features of traits such as commutative composition, explicit
conflict detection and resolution. In addition, our language also
supports instantiating and composing traits dynamically, which cannot
be achieved in many existing statically typed object-oriented
languages.

In summary, the contributions of this paper are:

\begin{itemize*}

\item \textbf{Disjoint Intersection Types:} A new form of intersection type
where only disjoint types are allowed. A sound and complete algorithmic
specification of disjointness (with respect to the corresponding formal
definition) is presented.

\item \textbf{Dynamically Composable Traits:} As an application of this work, we show how to build a
trait-based object-oriented language by leveraging the type system of \name. The
resulting source language enjoys many features described in the original traits
proposal for free.

\item \textbf{Formalization of \name and Proof of Coherence:} An
   elaboration semantics of \name into $\lambda$-calculus is
   given. Type-soundness and coherence are proved and formalized. Several key theorems
   are mechanically formalized using the Coq theorem prover.

\item \textbf{Implementation:} An implementation of an extension of \name, as
well as encodings of the examples presented in the paper, are publicly available
online\footnote{
\url{https://github.com/hkuplg/fcore/tree/develop/examples/traits}}.

\end{itemize*}

\bruno{Records and record operations missing in the appendix.}

\begin{comment}

There has been a remarkable number of works aimed at improving support
for extensibility in programming languages. The motivation behind this
line of work is simple, and it is captured quite elegantly by the
infamous \emph{Expression Problem}~\cite{wadler1998expression}: there
are \emph{two} common and desirable forms of extensibility, but most
mainstream languages can only support one form well. Unfortunately
the lack of support in the other form has significant
consequences in terms of code maintenance and software evolution.  As a
result researchers proposed various approaches to address the problem,
including: visions of new programming
models~\cite{Prehofer97,Tarr99ndegrees,Harrison93subject}; new
programming languages or language
extensions~\cite{McDirmid01Jiazzi,Aracic06CaesarJ,Smaragdakis98mixin,nystrom2006j},
and \emph{design patterns} that can be used with existing mainstream
languages~\cite{togersen:2004,Zenger-Odersky2005,oliveira09modular,oliveira2012extensibility}.

Some of the more recent work on extensibility is focused on design
patterns. Examples include \emph{Object
  Algebras}~\cite{oliveira2012extensibility}, \emph{Modular Visitors}~\cite{oliveira09modular,togersen:2004} or
Torgersen's~\cite{togersen:2004} four design patterns using generics. In those
approaches the idea is to use some advanced (but already available)
features, such as \emph{generics}~\cite{Bracha98making}, in combination with conventional
OOP features to model more extensible designs.  Those designs work in
modern OOP languages such as Java, C\#, or Scala.

Although such design patterns give practical benefits in terms of
extensibility, they also expose limitations in existing mainstream OOP
languages. In particular there are three pressing limitations:
1) lack of good mechanisms for
  \emph{object-level} composition; 2) \emph{conflation of
    (type) inheritance with subtyping}; 3) \emph{heavy reliance on generics}.

  The first limitation shows up, for example, in encodings of Feature-Oriented
  Programming~\cite{Prehofer97} or Attribute Grammars~\cite{Knuth1968} using Object
  Algebras~\cite{oliveira2013feature,rendel14attributes}. These programs are best
  expressed using a form of \emph{type-safe}, \emph{dynamic},
  \emph{delegation}-based composition. Although such form of
  composition can be encoded in languages like Scala, it requires the
  use of low-level reflection techniques, such as dynamic proxies,
  reflection or other forms of meta-programming. It is clear
  that better language support would be desirable.

  The second limitation shows up in designs for modelling
  modular or extensible visitors~\cite{togersen:2004,oliveira09modular}.  The vast majority of modern
  OOP languages combines type inheritance and subtyping.
  That is, a type extension induces a subtype. However
  as Cook et al.~\cite{cook1989inheritance} famously argued there are programs where
  ``\emph{subtyping is not inheritance}''. Interestingly
  not many programs have been previously reported in the literature
  where the distinction between subtyping and inheritance is
  relevant in practice. However, as shown in this paper, it turns out that this
  difference does show up in practice when designing modular
  (extensible) visitors.  We believe that modular visitors provide a
  compelling example where inheritance and subtyping should
  not be conflated!

  Finally, the third limitation is prevalent in many extensible
  designs~\cite{togersen:2004,Zenger-Odersky2005,oliveira09modular,oliveira2013feature,rendel14attributes}.
  Such designs rely on advanced features of generics,
  such as \emph{F-bounded polymorphism}~\cite{Canning89f-bounded}, \emph{variance
    annotations}~\cite{Igarashi06variant}, \emph{wildcards}~\cite{Torgersen04wildcards} and/or \emph{higher-kinded
    types}~\cite{Moors08generics} to achieve type-safety. Sadly, the amount of
  type-annotations, combined with the lack of understanding of these
  features, usually deters programmers from using such designs.

This paper presents \name (pronounced \emph{f-and}): an extension of
$\lambda$-calculus~\cite{Reynolds74f} with intersection types and a merge
operator~\cite{dunfield2014elaborating}.  The goal of \name is to study the
\emph{minimal} foundational language constructs that are needed to support
various extensible designs, while at the same time addressing the limitations of
existing OOP languages. To address the lack of good object-level composition
mechanisms, \name uses the merge operator for dynamic composition of
values/objects. Moreover, in \name (type-level) extension is independent of
subtyping, and it is possible for an extension to be a supertype of a base
object type. Furthermore, intersection types and conventional subtyping can be
used in many cases instead of advanced features of generics. Indeed this paper
shows many previous designs in the literature can be encoded without such
advanced features of generics.


Technically speaking \name is mainly inspired by the work of
Dunfield~\cite{dunfield2014elaborating}. Dunfield showed how to model a simply typed
calculus with intersection types and a merge operator. The presence of
a merge operator adds significant expressiveness to the language,
allowing encodings for many other language constructs as syntactic
sugar. \name differs from Dunfield's work in a few
as
the elaboration semantics into $\lambda$-calculus is done directly from the
source calculus with subtyping.
% In contrast, Dunfield has an additional step which eliminates subtyping.
% Secondly, a non-technical difference
% is that \name is aimed at studying issues of OOP languages and
% extensibility, whereas Dunfield's work was aimed at Functional
% Programming and he did not consider application to extensibility.
% Like many other foundational formal models for OOP (for
% example $F_{<:}$~\cite{CMMS}), \name is purely functional and it uses
% structural typing.

%%\name is
%%formalized and implemented. Furthermore the paper illustrates how
%%various extensible designs can be encoded in \name.

In summary, the contributions of this paper are:

\begin{itemize*}

\item {\bf A Minimal Core Language for Extensibility:} This paper
  identifies a minimal core language, \name, capable of
  expressing various extensibility designs in the literature.
  \name also addresses limitations of existing OOP
  languages that complicate extensible designs.

\item {\bf Formalization of \name:} An elaboration semantics of
  \name into $\lambda$-calculus is given, and type-soundness is proved.

\item {\bf Encodings of Extensible Designs:} Various encodings of
  extensible designs into \name, including \emph{Object
    Algebras} and \emph{Modular Visitors}.

\item {\bf A Practical Example where ``Inheritance is not Subtyping''
    Matters:} This paper shows that modular/extensible visitors
  suffer from the ``inheritance is not subtyping problem''.
%% Moreover
%% with extensible visitors the extension should become a
%% \emph{supertype}, not a subtype. \bruno{extension with accept method}

\item {\bf Implementation:} An implementation of an
  extension of \name, as well as the examples presented in the
  paper, are publicly available\footnote{{\bf Note to reviewers:} The code (and some machine
    checked proofs) can be found in the repository.}.

\end{itemize*}

\end{comment}

% \section{Introduction}

% Dunfield's work showed how many language features can be encoded in terms
% of intersection types with a merge operator. However two important
% questions were left open by Dunfield:

% \begin{enumerate}
% \item How to allow coherent programs only?

% \item If a restriction that allows coherent programs is in place, can
%   all coherent programs conform to the restriction?
% \end{enumerate}

% In other words question 1) asks whether we can find sufficient
% conditions to guarantee coherency; whereas question 2) asks
% whether those conditions are also necessary. In terms of technical
% lemmas that would correspond to:

% \begin{enumerate}

% \item Coherency theorem: $\Gamma \turns e : A \leadsto E_1 \wedge
%   \Gamma \turns e : A \leadsto E_2~\to~E_1 = E_2$.

% \item Completness of Coherency: ($\Gamma \turns_{old} e : A \leadsto E_1 \wedge
%   \Gamma \turns_{old} e : A \leadsto E_2~\to~E_1 = E_2) \to \Gamma
%   \turns e : A$.

% \end{enumerate}

% For these theorems we assume two type systems. On liberal type system
% that ensures type-safety, but not coherence ($\Gamma \turns_{old} e :
% A$); and another one that is both type-safe and coherent  ($\Gamma \turns e :
% A$). What needs to be shown for completness is that if a coherent
% program type-checks in the liberal type system, then it also
% type-checks in the restricted system.


% \special{papersize=8.5in,11in}
% \setlength{\pdfpageheight}{\paperheight}
% \setlength{\pdfpagewidth}{\paperwidth}

% \title{\name}

% \subsection{``Testsuite'' of examples}

% \begin{enumerate}

% \item $\lambda (x : Int * Int). (\lambda (z : Int) . z)~x$: This
%   example should not type-check because it leads to an ambigous choice
%   in the body of the lambda. In the current system the well-formedness
%   checks forbid such example.

% \item $\Lambda A.\Lambda B.\lambda (x:A).\lambda (y:B). (\lambda (z:A)
%   . z) (x,,y)$: This example should not type-check because it is not
%   guaranteed that the instantiation of A and B produces a well-formed
%   type. The TyMerge rule forbids it with the disjointness check.

% \item $\Lambda A.\Lambda B * A.\lambda (x:A).\lambda (y:B). (\lambda
%   (z:A) . z) (x,,y)$: This example should type-check because B is
%   guaranteed to be disjoint with A. Therefore instantiation should
%   produce a well-formed type.

% \item $(\lambda (z:Int) . z) ((1,,'c'),,(2,False))$: This example
%   should not type-check, since it leads to an ambigous lookup of
%   integers (can either be 1 or 2). The definition of disjointness is
%   crutial to prevent this example from type-checking. When
%   type-checking the large merge, the disjointness predicate will
%   detect that more than one integer exists in the merge.

% \item $(\lambda (f: Int \to Int \& Bool) . \lambda (g : Int \to Char \& Bool) . ((f,,g) : Int \to Bool)$:
%   This example
%   should not type-check, since it leads to an ambigous lookup of
%   functions. It shows that in order to check disjointness
%   of functions we must also check disjointness of the subcomponents.

% \item $(\lambda (f: Int \to Int) . \lambda (g : Bool \to Int) . ((f,,g) : Bool \& Int \to Int)$:
%   This example shows that whenever the return types overlap, so does the function type:
%   we can always find a common subtype for the argument types.
% \end{enumerate}

% \subsection{Achieving coherence}

% The crutial challenge lies in the generation of coercions, which can lead
% to different results due to multiple possible choices in the rules that
% can be used. In particular the rules subinter1 and subinter2 overlap and
% can result in coercions that are not equivalent. A simple example is:

% $(\lambda (x:Int) . x) (1,,2)$

% The result of this program can be either 1 or 2 depending on whether
% we chose subinter1 or subinter2.

% Therefore the challenge of coherence lies in ensuring that, for any given
% types A and B, the result of $A <: B$ always leads to the same (or semantically
% equivalent) coercions.

% It is clear that, in general, the following does not hold:

% $if~A <: B \leadsto C1~and~A <: B \leadsto C2~then~C1 = C2$

% We can see this with the example above. There are two possible coercions:\\

% \noindent $(Int\&Int) <: Int \leadsto \lambda (x,y). x$\\
% $(Int\&Int) <: Int \leadsto \lambda (x,y). y$\\

% However $\lambda (x,y). x$ and $\lambda (x,y). y$ are not semantically equivalent.

% One simple observation is that the use of the subtyping relation on the
% example uses an ill-formed type ($Int\&Int$). Since the type system can prevent
% such bad uses of ill-formed types, it could be that if we only allow well-formed
% types then the uses of the subtyping relation do produce equivalent coercions.
% Therefore the we postulate the following conjecture:

% $if~A <: B \leadsto C1~and~A <: B \leadsto C2~and~A, B~well~formed~then~C1 = C2$

% If the following conjecture does hold then it should be easy to prove that
% the translation is coherent.

% % \begin{mathpar}
% %   \inferrule
% %   {}
% %   {\jtype \epsilon {1 \mergeop 2} {\constraints {\code{Int} \disjoint \code{Int}} \code{Int} \inter \code{Int}}}
% % \end{mathpar}

% % \begin{definition}{(Disjointness)}
% % Two sets $S$ and $T$ are \emph{disjoint} if there does not exist an element $x$, such that $x \in S$ and $x \in T$.
% % \end{definition}

% % \begin{definition}{(Disjointness)}
% % Two types $A$ and $B$ are \emph{disjoint} if there does not exist an term $e$, which is not a merge, such that $\jtype \epsilon e A'$, $\jtype \epsilon e B'$, $A' \subtype A$, and $B' \subtype B$.
% % \end{definition}

% \section{Introduction}

% The benefit of a merge, compared to a pair, is that you don't need to explicitly extract an item out. For example, \lstinline@fst (1,'c')@

% \begin{definition}{Determinism}
% If $e : A_1 \hookrightarrow E_1$ and $e : A_2 \hookrightarrow E_2$,
% then $A_1 = A_2$ and $E_1 = E_2$.
% \end{definition}

% \emph{Coherence} is a property about the relation between syntax and semantics. We say a semantics is \emph{coherent} if the syntax of a term uniquely determines its semantics.

% \begin{definition}{Coherence}
% If $e_1 : A_1 \hookrightarrow E_1$ and $e_2 : A_2 \hookrightarrow E_2$,
% $E_1 \Downarrow v_1$ and $E_2 \Downarrow v_2$,
% then $v_1 = v_2$.
% \end{definition}

% \begin{definition}{Disjointness}
% Two types $A$ and $B$ are \emph{disjoint} (written as ``$\disjoint A B$'') if there does not exist a type $C$ such that $C \subtype A$ and $C \subtype B$ and $C \subtype A \inter B$.
% \end{definition}
