\section{Algorithmic Disjointness} \label{sec:alg-dis}

As promised, in this section we present the set of rules for determining whether
two types are disjoint, and more importantly, how to derive them from the
definition we have in the previous sections. The derived set of rules for
disjointness is proved to be sound and complete with respect to the definition.

\subsection{Derivation}

In this subsection, we illustrate how to derive the algorithmic disjointness
rules by showing a detailed example for functions. For the ease of discussion,
first we introduce some notations.

\begin{definition}[Set of common supertypes]
  For any two types $A$ and $B$, we can denote the \emph{set of their common
  supertypes} by \[ \commonsuper(A,B) \] In other words, a type $C \in \;
  \commonsuper(A,B)$ if and only  if $A \subtype C$ and $B \subtype C$.
\end{definition}

\begin{example}
  $\commonsuper(\tyint,\tychar)$ is empty, since $\tyint$ and $\tychar$
  share no common supertype.
\end{example}

Parellel to the notion of the set of common supertypes is the notion of the set
of common subtypes.

\begin{definition}[Set of common subtypes]
  For any two types $A$ and $B$, we can denote the \emph{set of their common
  subtypes} by \[ \commonsub(A,B) \] In other words, a type $C \in \; \commonsub(A,B)$
  if and only  if $C \subtype A$ and $C \subtype B$.
\end{definition}

\begin{example}
  $\commonsub(\tyint,\tychar)$ is an infinite set which contains $\tyint \inter
  \tychar$, $\tychar \inter \tyint$, $(\tyint \inter \tybool) \inter \tychar$
  and so on. But the type $\tybool$ is not inside, since it is not a subtype of
  $\tyint$.
\end{example}

\paragraph{Shorthand notation} For brevity, we will use \[ \mathcal{A} \to
\mathcal{B} \] as a shorthand for the \emph{set} of types of the form $A \to B$,
where $A \in \mathcal{A}$ and $B \in \mathcal{B}$. The same shorthand applies to
all other constructors of types, in addition to functions, as well. As a simple
example,  \[ \{ \tyint, \tystring \} \to \{ \tyint, \tychar \} \] is a shorthand for \[ \{
\tyint \to \tyint, \tyint \to \tychar, \tystring \to \tyint, \tystring \to \tychar \} \]

Recall that we say two types $A$ and $B$ are disjoint if they do not share a
common supertype. Therefore, determining if two types $A$ and $B$ are disjoint
is the same as determining if $\commonsuper(A,B)$ is empty.

\paragraph{Determining disjointness of functions} Now let's dive into the case
where both $A$ and $B$ are functions and consider how to compute
$\commonsuper(A_1 \to A_2, B_1 \to B_2)$. By the subtyping rules, the supertype
of a function must also be a function.\george{Nah... only after normalizatoin.
If not, it can also be $\inter$} Let $C_1 \to C_2$ be a common supertype
of $A_1 \to A_2$ and $B_1 \to B_2$. Then it must satisfy the following:
\begin{mathpar}
  \inferrule
    {C_1 \subtype A_1 \\ A_2 \subtype C_2}
    {A_1 \to A_2 \subtype C_1 \to C_2}

  \inferrule
    {C_1 \subtype B_1 \\ B_2 \subtype C_2}
    {B_1 \to B_2 \subtype C_1 \to C_2}
\end{mathpar}
From which we see that $C_1$ is a common subtype of $A_1$ and $B_1$ and that
$C_2$ is a common supertype of $A_2$ and $B_2$. Therefore, we can wrtie:
\[ \commonsuper(A_1 \to A_2, B_1 \to B_2) \ = \ \commonsub(A_1,B_1) \to \commonsuper(A_2,B_2) \]
By definition, $\commonsub(A_1,B_1) \to \commonsuper(A_2,B_2)$ is not empty if and only if both
$\commonsub(A_1,B_1)$ and $\commonsuper(A_2,B_2)$ is nonempty. However, note
that with intersection types, $\commonsub(A_1,B_1)$ is always nonempty because
$A_1 \inter B_1$ belongs to $\commonsub(A_1,B_1)$. Therefore, the problem of
determining if $\commonsuper(A_1 \to A_2, B_1 \to B_2)$ is empty reduces to the
problem of determining if $\commonsuper(B_1 \to B_2)$ is empty, which is, by
definition, if $B_1$ and $B_2$ are disjoint. Finally, we have derived a rule for
functions:
\begin{mathpar}
  \ruledisfun
\end{mathpar}

The analysis needed for determining if types with other constructors are
disjoint is similar. Below are the major results of the recursive definitions for
$\commonsuper$:
\begin{align*}
  \commonsuper(A_1 \to A_2, B_1 \to B_2) &= \commonsub(A_1,B_1) \to \commonsuper(A_2,B_2) \\
  \commonsuper({A_1 \inter A_2, B})      &= \commonsuper(A_1, B) \cup \commonsuper(A_1,B) \\
  \commonsuper({A, B_1 \inter B_2})      &= \commonsuper(A, B_1) \cup \commonsuper(A,B_2)
\end{align*}\george{Missing the forall case. But we're just going to drop the formulae.}

\subsection{Rules}

The rules for the disjointness judgement are shown in
Figure~\ref{fig:disjointness}, which consists of two judgements.

\paragraph{Main judgement} The judgement $\jdisimpl \Gamma A B$ says two types
$A$ and $B$ are disjoint in a context $\Gamma$. \reflabel{\labeldisvar} is the
base rule and \reflabel{\labeldissym} is its twin rule.
\reflabel{\labeldisinterl} and \reflabel{\labeldisinterr} inductively distribute
the relation itself over the intersection constructor ($\inter$).
\reflabel{\labeldisfun} is the most interesting rule---it says two function
types are disjoint if and only if their return types are disjoint (regardless of
their parameter types!). For example, consider the two function types:
\[ \tyint \to \tystring \qquad \tybool \to \tystring \]
Even though their parameter types are disjoint, we are still able to think of a
type which is a supertype for both of them. For example, $ \tyint \inter \tybool
\to \tystring $. Therefore, by definition, the two function types are not
disjoint. The lesson from this example is that the parameter types of two
function types do not have a voice in determining whether that two function
types are disjoint or not, because in our system given any two types there
exists a subtype of the both.

\reflabel{\labeldisax} says two types are considered disjoint if they are
judged to be disjoint by the axiom rules, which is explained below.

\paragraph{Axioms} Up till now, the rules of $ \jdisimpl \Gamma A B $ have only
taken care of two types of the same constructors. But how can be the fact that
$\tyint$ and $\tyint \to \tyint$ are disjoint be decided? That is exactly the
place where the judgement $ A \disjointax B $ comes in handy. It provides the
axioms for disjointness. What is captured by the set of rules is that $ A
\disjointax B $ holds for all two types of different constructors unless any of
them is an intersection type. That is because for example, $ \tyint \inter
(\tychar \to \tychar) $ and $ \tychar \to \tychar $ have different constructors
and yet are not disjoint. \george{Can you decide that $\bot \disjoint \tyint \inter \tychar$?}

\begin{figure*}[h]
  \begin{mathpar}
    \formdis \\
    \ruledisvar \and \ruledissym \and \ruledisinterl \and \ruledisinterr \and
    \ruledisfun \and \ruledisforall \and \ruledisatomic
  \end{mathpar}

  \begin{mathpar}
    \formax \\
    \ruleaxbotfun \and \ruleaxbotforall \and \ruleaxfunforall \and \ruleaxsym
  \end{mathpar}

  \caption{Algorithmic Disjointness.}
  \label{fig:disjointness}
\end{figure*}

\subsection{Metatheory}

The algorithmic rules for disjointness is sound and complete.

\begin{lemma}{Symmetry of disjointness} \label{lemma:symmetry-of-disjointness}
  If $\jdis \Gamma A B$, then $\jdis \Gamma B A$.
\end{lemma}

\begin{theorem} \label{lemma:disjoint-intersect}
  If $\jdis \Gamma A C$ and $\jdis \Gamma B C$,
  then $\jdis \Gamma {A \inter B} {C}$.
\end{theorem}

\begin{lemma} \label{lemma:common-supertype}
  If $A_1 \to A_2 \subtype D$ and $B_1 \to B_2 \subtype D$,
  then there exists a $C$ such that $A_2 \subtype C$ and $B_2 \subtype C$.
\end{lemma}

\begin{restatable}[Soundness]{theorem}{soundness}
  \label{theorem:soundness}

  For any two types $A$, $B$, $\jdisimpl \Gamma A B$ implies $\jdis \Gamma A B$.
\end{restatable}

\begin{proof}
  By induction on the derivation of $\jdisimpl \Gamma A B$.
\end{proof}

\begin{restatable}[Completeness]{theorem}{completeness}
  \label{theorem:completeness}

  For any two types $A$, $B$, $\jdis \Gamma A B$ implies $\jdisimpl \Gamma A B$.
\end{restatable}

\begin{proof}
  By induction on the derivation of $\jdis \Gamma A B$.
\end{proof}
