\section{Overview} \label{sec:overview}

\subsection{Motivating Design Choices}

We need to motivate the 3 changes:

\paragraph{Well-formed types}: We need a new notion of well-formed types.

\paragraph{Disjoint quantification}: Needed when we have a
combination of polymorphism and intersection types.

With a subtyping relation in a type system, bounded polymorphism extends the universal quantifier by confining the polymorphic type to be a subtype of a given type. In our type system, the forall binder also extends the parametric polymorphism, but in a different vein: the polymorphic type can only be disjoint with a given type. Later during an instantiation, if the type provided overlaps with the constraint, such instantiation will be rejected by our type system.

\begin{itemize}
  \item \textbf{Bounded polymorphism}---the instantiation can only be the descendant of a given type
  \item \textbf{Polymorphism with disjoint constraint}---the instantiation cannot share a common ancestor with a given type
\end{itemize}

% \bruno{mention (and cite) that there are a number of OO languages supporting
% intersection types }
%\george{Change the examples later to something very simple.}
%\bruno{Mention that we use some syntactic sugar, which is part
%of our implementation.}

% \bruno{Syntax in the examples is not being used consistently!
% That is why it is better to import code from a file. You have to
% carefully go over every example and see whether the examples
% are actually valid syntax in our language. I fixed some of these,
% but there may be some more.}

This section provides the reader with the necessary intuition for
\name by informally introducing the main features of the language.
The features of \name are also contrasted with features of mainstream
languages such as Java or Scala. Note that this section uses some
common syntactic sugar in programming languages, which is
part of our implementation.

%will do so by introducing a simple source language, which will be used in
%Section~\ref{sec:overview}. \george{From the practice of similar papers, I
% think a figure listing the source syntax is necessary.}

\subsection{Intersection types in existing languages}\label{subsec:interScala}

A number of OO languages, such as Java, C\#, Scala, and
Ceylon\footnote{\url{http://ceylon-lang.org/}}, already support intersection
types to different degrees. In Java, for example,

\begin{lstlisting}
interface AwithB extends A, B {}
\end{lstlisting}

\noindent introduces a new interface \lstinline{AwithB} that satisfies the interfaces of
both \lstinline{A} and \lstinline{B}. Arguably such type can be considered as a nominal
intersection type. Scala takes one step further by eliminating the
need of a nominal type. For example, given two concrete traits, it is possible to
use \emph{mixin composition} to create an object that implements both
traits. Such an object has a (structural) intersection type:

\begin{lstlisting}
trait A
trait B

val newAB : A with B = new A with B
\end{lstlisting}

\noindent Scala also allows intersection of type parameters. For example:
\begin{lstlisting}
def merge[A,B] (x: A) (y: B) : A with B = ...
\end{lstlisting}
uses the annonymous intersection of two type parameters \lstinline{A} and
\lstinline{B}. However, in Scala it is not possible to dynamically
compose two objects. For example, the following code:

\begin{lstlisting}
// Invalid Scala code:
def merge[A,B] (x: A) (y: B) : A with B = x with y
\end{lstlisting}

\noindent is rejected by the Scala compiler. The problem is that the
\lstinline{with} construct for Scala terms can only be used to
mixin traits or classes, and not arbitrary objects. Note that in the
definition \lstinline{newAB} both \lstinline{A} and \lstinline{B} are
\emph{traits}, whereas in the definition of \lstinline{merge} the variables
\lstinline{x} and \lstinline{y} denote \emph{objects}.

\begin{comment}
A common limitation in those languages, though, is that there is no introduction
construct at the term level for intersection types. In Java and Scala, we cannot
create an instance of class type \lstinline{A & B} with
\begin{lstlisting}
  new A() & B()
\end{lstlisting}
\end{comment}

This limitation essentially put intersection types in Scala in a second-class
status. Although \lstinline{merge} returns an intersection type, it is
hard to actually build values with such types. In essense an
object-level introduction contruct for intersection types is missing.
As it turns out using low-level type-unsafe programming features such
as dynamic proxies, reflection or other meta-programming techniques,
it is possible to implement such an introduction
construct in Scala~\cite{oliveira2013feature,rendel14attributes}. However, this
is clearly a hack and it would be better to provide proper language
support for such a feature.

\begin{comment}
This is in
contrast, there are term-level introduction construct for function types (with
lambdas) and universal quantification (with big lambdas) in most core
calculi.
\end{comment}

% The key constructs are the ``merge'' operator, denoted by $ \mergeop $
% at the term level and the corresponding type intersection operator, denoted by
% $ \inter $ at the type level.
% \bruno{We take a particular approach to intersection types, namely that of
% Dunfield. Some other approaches are different. The \emph{key feature} in
% Dunfield's approach is the $\mergeop$ operator, which allows for run-time
% value/object composition.}

% The addition of intersection types to System $ F $ has a number of consequences,
% which we will explore one by one in the following subsections.
% \bruno{Don't mention prototype-based inheritance in the summary, unless you
% are going to mention it later in the section. \emph{More generally don't
% mention things that you don't talk about later!}}.

\subsection{Intersection types in \name}

% What is an intersection type? One classic view is from set-theoretic
% interpretation of types: $ A \inter B $ stands for the intersection of
% the set of values of $ A $ and $ B $. A more practical view, adopted in this
% paper, regards types as a kind of interface: a value of type
% $ A \inter B $ satisfies both of the interfaces of $ A $ and $ B $. For
% example, \lstinline{eval : Int} is the interface that supports evaluation to
% integers, while \lstinline{eval : Int & print : String } supports both
% evaluation and pretty printing. Those interfaces are akin to interfaces in
% Java or traits in Scala. But one key difference is that they are unnamed in
% \name.

To address the limitations of intersection types in languages like
Scala, \name allows intersecting any two terms at run time using a
\emph{merge} operator (denoted by $ \mergeop $)~\cite{dunfield2014elaborating}.  With the merge
operator it is trivial to implement the \lstinline{merge} function in \name:

\begin{lstlisting}
let merge[A,B] (x : A) (y : B) : A & B = x ,, y;
\end{lstlisting}

\noindent In contrast to Scala's term-level \lstinline{with}
construct, the operator \lstinline{,,} allows two arbitrary values \lstinline{x}
and \lstinline{y} to be merged. The resulting type is an
intersection of the types of  \lstinline{x}
and \lstinline{y} (\lstinline{A & B} in this case).

\begin{comment}
The following table
summarizes the extent of support for intersection types in Java,
Scala, and \name.

\hspace{-13pt}\begin{tabular}{ l | c | c | c  }
                                   & Java       & Scala      & \name      \\ \hline
  Basic intersection types         & \checkmark & \checkmark & \checkmark \\ \hline
  Anonymous intersection types     &            & \checkmark & \checkmark \\ \hline
  Intersection of type parameters  &            & \checkmark & \checkmark \\ \hline
  Term-level intersection          &            &            & \checkmark
\end{tabular} \\
\end{comment}

%In \name the central addition to the type system of System $F$ are
%intersection types.

\paragraph{Intersection types for overloading.}
A typical use-case for intersection types is to do
\emph{overloading}. The benefit is that programmers can use the same
operation on different types and delegate the task of choosing a
concrete implementation to the type system. For example, we can define
a \lstinline{show} function that takes either an integer or a boolean
and returns its string representation. In other words, \lstinline{show} is also
\emph{both} a function from integers to strings as well as a function
from boolean to strings.  Therefore, in \name \lstinline{show} should be of
following type:
\begin{lstlisting}
show : (Int -> String) & (Bool -> String)
\end{lstlisting}
Assuming that the following two functions are available:
\begin{lstlisting}
showInt : Int  -> String
showBool : Bool -> String
\end{lstlisting}

\noindent The overloaded \lstinline{show} function is defined by
merging the \lstinline{showInt} and \lstinline{showBool} using the
merge operator:

\begin{lstlisting}
let show = showInt ,, showBool;
\end{lstlisting}
% \bruno{This is not a proper sentence. Polish text.}

To illustrate the usage, consider the function application \lstinline{show 100}.
The type system will pick the first component of \lstinline{show}, namely
\lstinline{showInt}, as the implementation being applied to \lstinline{100}
because the type of \lstinline{showInt} is compatible with \lstinline{100}, but
\lstinline{showBool} is not. This example shows that one may regard
intersections in our system as ``implicit pairs'' whose introduction is explicit
by the merge operator and elimination is implicit (with no source-level
construct for elimination).

% The merge construct in the original function is elaborated into a pair in the
% target language:

% \begin{verbatim}
% show = (showInt, showBool)
% \end{verbatim}

% In the target language where there is no intersection types, the application
% of the integer \texttt{1} to this function does not typecheck. However, we may
% rescue this situtation by inserting a coercion that extracts the first item
% out of this pair.

% Thus \texttt{show 1} in FI corresponds to \texttt{(fst show) 1} in F.

% While elaborating intersection types, this paper is the first that presents a
% type system that incorporates both parametric polymorphism and intersection
% polymorphism.

% Describe intersection types, encoding records with Intersecion types

% \lstinputlisting[linerange=-]{} % APPLY:linerange=MIXIN_LIB

\paragraph{Subtyping.} The previous example exploits a natural
subtyping relation on intersection types. That is the type

\begin{lstlisting}
(Int -> String) & (Bool -> String)
\end{lstlisting}

\noindent is a \emph{subtype} of both \lstinline{Int -> String} and
\lstinline{Bool -> String}. This is why \lstinline{show} can take
\lstinline{100} as its argument. Generally speaking an intersection type
\lstinline{A & B} is a subtype of both \lstinline{A} and \lstinline{B}.
Moreover, subtyping of intersection types in \name is purely structural
and it enjoys properties such as \emph{idempotence},
\emph{commutativity} and \emph{associativity}
(equality is defined as bidirectional subtyping relation): \\

% \noindent {\bf commutativity: } $A \inter B = B \inter A$  \\
% {\bf associativity:~~~~~~}$A \inter (B \inter C) = (A \inter B) \inter C$

\begin{tabular}{llrcl}
  \textbf{Idempotent}  & $A \inter A$            & $=$ & $A$ \\
  \textbf{Commutative} & $A \inter B$            & $=$ & $B \inter A$ \\
  \textbf{Associative} & $(A \inter B) \inter C$ & $=$ & $A \inter (B \inter C)$
\end{tabular}

% \bruno{I think this will be a good place to talk about the following material:}
% \bruno{Talk about Inheritance is not Subtyping.
% Describe type inheritance and subtyping, show that they don't necessarelly
% go along together in our language. You may need to write some Java code, to
% illustrate differences. \emph{We support contravariant argument types!}
% }
% \bruno{Related to the previous point, don't forget to mention that
% there are nominal languages, that also separate inheritance from subtyping!
% See Klaus Ostermann's paper \& ``Inheritance is not Subtyping''.
% }

\paragraph{Type-inheritance} In \name type-inheritance is decoupled
from subtyping. Type-inheritance, as formally defined by Cook et
al.~\cite{cook1989inheritance}, is:
\[
G = F + \{ l_1:A_1, \ldots, l_n:A_n \}
\]
Here $G$ is a new type defined by adding fields to an existing type $F$.
The argument that ``Inheritance is not subtyping'' means that structually, $G$ can
be: 1) a subtype of $F$, 2) a supertype of $F$, or 3) $G$ and $F$ can have no
relation at all! If none of $l_1, \ldots, l_n$ is present in $F$, $G$ can be
safely regarded as a subtype of $F$, since $G$ always contains the fields
required by the signature $F$; if all of $l_1, \ldots, l_n$ are present in $F$
and each of their types is a supertype of the corresponding type, $G$ becomes
instead a supertype of $F$; Conceivably, the third possibility will make $F$ and
$G$ have no relation. However, most OO languages conflate the
subtyping and type-inheritance, allowing only 1). This leads
to cases where the type system prevents expressing semantically
correct programs. For example, consider a program that consists of the two
types:
\begin{lstlisting}
type A = {x1: Int};
type F = {f: A -> Int};
\end{lstlisting}
Later one wishes to add a field to form a new type \lstinline$B$ and refine the
parameter type of \lstinline$f$ by using \lstinline$B$ instead:
\begin{lstlisting}
type B = {x1: Int, x2: Int}; -- B <: A
type G = {f: B -> Int};      -- G :> F
\end{lstlisting}
Clearly, \lstinline$B$ should be a subtype of \lstinline$A$. But it is also
desirable to have the property that \lstinline$G$ is a \emph{supertype} of
\lstinline$F$, since the refinement happens at a parameter position. In OO
languages such as Java and Scala, the idiom of making \lstinline$G$ type-inherit
$F$ is through \lstinline$G extends F$, which only works when $G$ is a
subtype of $F$.
%Unfortunately, $G$ has become a subtype
%of $F$.

Although in nominal systems there is the well-known tension between inheritance
and subtyping, it is worth noting that extending this benefit to nominal
subtyping is possible. Ostermann~\cite{ostermann2008nominal} proposed a flexible
nominal system that untangles inheritance and subtyping. For example, users may
declare supertype of an existing class type while at the same time inheriting
the implementation, and vice versa.

% \george{I think points regarding the designs of type system should prob. be addressed
%   somewhere else. The purpose of this section is just introducing the source
%   language and preparing the reader for the next section.}

% Subtyping in \name is syntactical and structural.

\subsection{\hspace{-5pt}Intersection~types~and~parametric~polymorphism}

The combination of intersection types and parametric polymorphism
makes System \name quite expressive. In particular this combination
enables an encoding of a simple form of \emph{bounded universal
  quantification}~\cite{cardelli1985understanding}.

%In the same way that System \name extends System $ F $ with
%intersection types, System $ F_{\subtype} $ extends System $ F $ with bounded
%polymorphism. $ F_{\subtype} $~\cite{pierce2002types} allows giving an
%upper bound to the type variable in type abstractions.

\paragraph{Bounded quantification and loss of information.}
The idea of bounded universal quantification was discussed in the
seminal paper by Cardelli and Wegner
~\cite{cardelli1985understanding}. They show that bounded quantifiers
are useful because they are able to solve the ``loss of information''
problem. The extension of System $ F $ with intersection types is able
to address the same problem effectively. Suppose we have the following
definitions:
\begin{lstlisting}{language=scala}
let user = {name = "George", admin = true};
let id(user: {name: String}) = user;
\end{lstlisting}
Under a structural type system, passing \lstinline{user}
to \lstinline{id} is allowed.
In other words, the type of \lstinline{user}
is of a subtype of the expected parameter type of \lstinline{id}.
However there is a problem: what if programmers
want to access the \texttt{admin} field later. For example:
\begin{lstlisting}{language=scala}
(id user).admin
\end{lstlisting}

They cannot do so as the above will not typecheck. After going through the
function, the resulting value has the type:

\begin{lstlisting}{language=scala}
{name: String}
\end{lstlisting}
This is rather undesired because the value does have an \texttt{admin} field!

Bounded polymorphism enables the \lstinline{id} function to return the exact type of the
argument so that there is no problem in accessing the \texttt{admin} field later.
Consider the example below:
\begin{lstlisting}{language=scala}
let id[A <: {name: String}] (user: A) = user;
(id [{name: String, admin: Bool}] user).admin
\end{lstlisting}

\noindent This piece of pseudo-code, which is not valid in \name due
to the use of bounded polymorphism, illustrates the idea. Instead of
giving the \lstinline{user} argument a concrete type, the
\lstinline{id} function specifies that such an argument is a subtype
of \{\lstinline{name: String}\}. With such a type the function
\lstinline{id} avoids losing information about the argument type.


\paragraph{Encoding bounded polymorphism in \name.}
\name does not have bounded polymorphism. However the same effect can
be achieved with a combination of intersection types and parametric polymorphism:
\begin{lstlisting}{language=F2J}
let id[A] (user: A & {name: String}) = user;
(id [{admin: Bool}] user).admin
\end{lstlisting}
By requiring the type of the argument to be an intersection type of a type
parameter and the upper bound and passing the type information, we make sure
that we can still access the \lstinline{admin} field later.

Therefore, this technique allows \name to encode a simple form
of bounded quantification. This is good because it means that \name
can express many common idioms that require bounded quantification
without complicating the core calculus with native support for bounded quantifiers.

% \cite{pierce1997intersection}

% \bruno{I don't think this has been shown. What we can say is: we can encode a
% form of bounded polymorphism with intersection types.}

\subsection{Applications to extensibility} \label{sec:overview}

% Structural subtyping facilitates reuse~\cite{malayeri2008integrating}.
% \bruno{orphan sentence!}

% \bruno{Make sure that the important code in the paper is reused from a script
%   and not inlined directly in the text.}

This section shows that, although \name is a minimal language, its
features are enough for encoding extensible designs that been
presented in mainstream languages. Moreover \name addresses
limitations of those languages, making those designs significantly
simpler. There are two main advantages of \name over existing
languages:

\begin{enumerate}
\item \name supports dynamic composition of intersecting values.
\item \name does not couple type inheritance and subtyping. Moreover
  \name supports contravariant parameter types in the subtyping relation.
\end{enumerate}

These two features avoid the use of low-level programming techniques,
and make the designs less reliant on advanced features of generics.

\begin{comment}
Various solutions have been proposed to deal with the extensibility problems and
many rely on heavyweight language features such as abstract methods and classes
in Java. These two features can be used to improve existing designs of modular
programs.

\bruno{I would like to see a story about Church Encodings in
  \name. Can you look at Pierce's papers and try to write something
  along those lines? That will be a good intro for object algebras and
visitors!}


\url{http://repository.cmu.edu/cgi/viewcontent.cgi?article=3018&context=compsci}
Church encoding allows modelling algebraic data types.
\end{comment}

% Introduce the expression problem



% There has been recently a lightweight solution to the expression problem that
% takes advantage of covariant return types in Java. We show that \name is able
% to solve the expression problem in the same spirit.

% - Object/Fold Algebras. How to support extensibility in an easier way.

% See Datatypes a la Carte

% - Mixins

% - Lenses? Can intersection types help with lenses? Perhaps making the
% types more natural and easy to understand/use?

% - Embedded DSLs? Extensibility in DSLs? Composing multiple DSL interpretations?

% http://www.cs.ox.ac.uk/jeremy.gibbons/publications/embedding.pdf

% \bruno{You already talk about overloading in the previous section. Need to
% decide where to put the text!}

% Dunfield~\cite{dunfield2014elaborating} notes that using merges as a mechanism
% of overloading is not as powerful as type classes.

% Multiple inheritance?
% Algebra -> P1,2
% Visitor -> P2

% Yanlin
% Mixin

% \begin{lstlisting}
% let merge A B (f : ExpAlg A) (g : ExpAlg B) = {
%   lit = \(x : Int). f.lit x ,, g.lit x,
%   add = \(x : A & B). \(y : A & B). f.add x y ,, g.add x y
% };
% \end{lstlisting}

\subsection{Object algebras}\label{subsec:OAs}

% Object algebras provide an alternative to \emph{algebraic data types}
% (ADT).\bruno{We are targeting an OO crowd. Mentioning algebraic
%   datatypes is not going to be very useful there.}

%  For example, the
% following Haskell definition of the type of simple expressions
% \begin{lstlisting}{language=haskell}
% data Exp where
%   Lit :: Int -> Exp
%   Add :: Exp -> Exp -> Exp
% \end{lstlisting}
% can be expressed by the \emph{interface} of an object algebra of
% simple expressions:
% \begin{lstlisting}{language=scala}
% trait ExpAlg[E] {
%   def lit(x: Int): E
%   def add(e1: E, e2: E): E
% }
% \end{lstlisting}
% Similar to ADT, data constructors in object algebras are represented by functions such as
% \lstinline{lit} and \lstinline{add} inside an interface \lstinline{ExpAlg}.
% Different with ADT, the type of the expression itself is abtracted by a type
% parameter \lstinline{E}.

% which can be expressed similarly in \name as:
% \begin{lstlisting}{language=F2J}
% type ExpAlg E = {
%   lit : Int -> E,
%   add : E -> E -> E
% }
% \end{lstlisting}

% Introduce Scala's intersection types

% Scala supports intersection types via the \lstinline{with} keyword. The type
% \lstinline{A with B} expresses the combined interface of \lstinline{A} and
% \lstinline{B}. The idea is similar to
% \begin{lstlisting}{language=java}
% interface AwithB extends A, B {}
% \end{lstlisting}
% in Java.
% \footnote{However, Java would require the \lstinline{A} and \lstinline{B} to be
%   concrete types, whereas in Scala, there is no such restriction.}

% The value level counterpart are functions of the type \lstinline
% {A => B => A with B}. \footnote{FIXME}

Oliveira and Cook~\cite{oliveira2012extensibility} proposed a design pattern that can solve the
Expression Problem in languages like Java. An advantage of the pattern
over previous solutions is that it is relatively lightweight in terms
of type system features. In a latter paper, Oliveira et al.~\cite{oliveira2013feature}
noted some limitations of the original design pattern and proposed
some new techniques that generalized the original pattern, allowing it
to express programs in a Feature-Oriented Programming~\cite{Prehofer97} style.
Key to these techniques was the ability to dynamically compose object
algebras.

Unfortunatelly, dynamic composition of object algebras is
non-trivial. At the type-level it is possible to express the resulting
type of the composition using intersection types. Thus, it is still
possible to solve that part problem nicely in a language like Scala (which
has basic support for intersection types). However, the dynamic
composition itself cannot be easily encoded in Scala. The fundamental
issue is that Scala lacks a \lstinline{merge} operator (see the
discussion in Section~\ref{subsec:interScala}). Although both Oliveira et al.~\cite{oliveira2013feature} and
Rendell et al.~\cite{rendel14attributes} have shown that such a \lstinline{merge} operator can
be encoded in Scala, the encoding fundamentally relies in low-level
programming techniques such as dynamic proxies, reflection or
meta-programming.

Because \name supports a \lstinline{merge} operator natively, dynamic
object algebra composition becomes easy to encode. The remainder of
this section shows how object algebras and object algebra composition
can be encoded in \name. We will illustrate this point
step-by-step by solving the Expression Problem.
%%Prior knowledge of object algebras is not assumed.

% can be cumbersome and
% language support for intersection types would solve that problem.
% Our type system is just a simple extension of System $ F $; yet surprisingly, it
% is able to solve the limitations of using object algebras in languages such as
% Java and Scala.

\paragraph{A simple system of arithmetic expressions.}
In the Expression Problem, the idea is to start with a very simple
system modeling arithmetic expressions and evaluation.
The initial system considers expressions with two variants (literals and
addition) and one operation (evaluation). Here is an interface that supports
evaluation:
\begin{comment}
  \begin{lstlisting}{language=F2J}
    type IEval = {eval: Int};
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=4-4]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_IEVAL

\noindent In \name the interfaces of objects (or object types) are expressed as
a record type. A \lstinline{type} declaration allow us to create a
simple alias for a type.  In this case \lstinline{IEval} is an alias
for \{\lstinline{eval: Int}\}.

With object algebras, the idea is to create an object algebra
interface, \lstinline$ExpAlg$, for expression types with the two
variants. This interface has a fixed number of variants, but abstracts over the
type of the interpretation \lstinline$E$.

\begin{comment}
  \begin{lstlisting}{language=F2J}
    type ExpAlg[E] = {
      lit: Int -> E,
      add: E -> E -> E
    };
  \end{lstlisting}
\end{comment}

\lstinputlisting[linerange=8-11]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_EXPALG

% whereas
%\lstinline$ExpAlg[IEval & IPrint]$ will be the type of object algebras that
%support both evaluation and pretty printing.
% In \name, record types are structural and hence any value that satisfies this
% interface is of type \lstinline$IEval$ or of a subtype of \lstinline$IEval$.
% \footnote{Should be mentioned in S2.}
Having defined the interfaces, we can implement that object algebra interface
with \lstinline$evalAlg$, which is an object algebra for evaluation.
\begin{comment}
  \begin{lstlisting}{language=F2J}
    let evalAlg: ExpAlg[IEval] = {
      lit = \(x: Int) -> {eval = x},
      add = \(x: IEval) (y: IEval) -> {
        eval = x.eval + y.eval
      }
    };
  \end{lstlisting}
\end{comment}

\lstinputlisting[linerange=15-20]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_EVALALG
In this example we implement a record, where the two operations
\lstinline{lit} and \lstinline{add} return a record with type \lstinline{IEval}.
The type \lstinline$ExpAlg[IEval]$ is the type of object algebras
supporting evaluation. However, the one interesting point
of object algebras is that other operations can be supported as
well.

\paragraph{Add a subtraction variant.} The point of the Expression
Problem is to support the addition of new features to the existing
program, without modifying existing code.
The first feature is adding a new variant, such as subtraction. We can do so by
simply intersecting the original types and merging with the original values:

\begin{comment}
  \begin{lstlisting}{language=F2J}
    type SubExpAlg[E] =
    ExpAlg[E] & {sub: E -> E -> E};
    let subEvalAlg = evalAlg ,, {
      sub = \(x: IEval) (y: IEval) -> {
        eval = x.eval - y.eval
      }
    };
  \end{lstlisting}
\end{comment}

\lstinputlisting[linerange=24-31]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_SUBEXPALG

\noindent Note that here intersection types are used to model \emph{type
  inheritance} and the merge operator models a basic form of
\emph{dynamic implementation inheritance}.

\paragraph{Add a pretty printing operation.}
A second extension is adding a new operation, such as pretty printing.
Similar to evaluation, the interface of the pretty printing feature
is modeled as:
\begin{comment}
  \begin{lstlisting}{language=F2J}
    type IPrint = {print : String};
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=35-35]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_IPRINT
The implementation of pretty printing for expressions that support literals,
addition, and subtraction is:
\begin{comment}
  \begin{lstlisting}{language=F2J}
    let printAlg : SubExpAlg[IPrint] = {
      lit = \(x: Int) -> {print = x.toString()},
      add = \(x: IPrint) (y: IPrint) -> {
        print = x.print ++ " + " ++ y.print
      },
      sub = \(x: IPrint) (y: IPrint) -> {
        print = x.print ++ " - " ++ y.print
      }
    };
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=39-47]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_PRINTALG

\paragraph{Usage.}
With the definitions above, values are created using the
appropriate algebras. For example, the expression \lstinline{7 + 2}
is encoded as follows:
\begin{comment}
  \begin{lstlisting}{language=F2J}
    let e1[E] (f: SubExpAlg[E]) =
    f.add (f.lit 7) (f.lit 2);
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=51-52]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_E1

\noindent The expressions are unusual in the sense that they are
functions that take an extra argument \lstinline$f$. The extra
argument is an object algebra that uses the functions in the record
(\lstinline$lit$, \lstinline$add$ and \lstinline$sub$) as factory
methods for creating values. Moreover, the algebras themselves are
abstracted over the allowed operations such as evaluation and pretty
printing by requiring the expression functions to take an extra
argument \lstinline$E$.

\paragraph{Dynamic object algebra composition.}
To obtain an expression that supports both evaluation and pretty
printing, a mechanism to combine the evaluation and printing
algebras is needed. \name allows such composition: the \lstinline$combine$
function, which takes two object algebras to create a combined algebra. It
does so by constructing a new object algebra where each field is a
function that delegates the input to the two algebra parameters.
\begin{comment}
  \begin{lstlisting}{language=F2J}
    let combine[A,B](f: ExpAlg[A])(g: ExpAlg[B]) :
    ExpAlg[A&B] = {
      lit = \(x: Int) -> f.lit x ,, g.lit x,
      add = \(x: A & B) (y: A & B) ->
      f.add x y ,, g.add x y
    }
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=58-63]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_COMBINE

\begin{comment}
  \begin{lstlisting}{language=F2J}
    let newAlg =
    combine[IEval,IPrint] subEvalAlg printAlg;
    let o = e1[IEval&IPrint] newAlg;
    o.print ++ " = " ++ o.eval.toString()
  \end{lstlisting}
\end{comment}

\lstinputlisting[linerange=67-69]{../src/ObjectAlgebra.sf} % APPLY:linerange=OA_USAGE

Note that \lstinline$o$ is a single object that supports both
evaluation and printing, as the output of the program is
\begin{lstlisting}
> 7 + 2 = 9
\end{lstlisting}

In contrast to the Scala solutions available in the
literature, \name is able to express object algebra
composition very directly by using the merge operator.

\subsection{Back to visitors}

Object Algebras are closely related to the visitor
pattern~\cite{gamma1994design}.  Indeed, object algebra interfaces are just
\emph{internal
  visitors}~\cite{oliveira09modular,oliveira2012extensibility}.
What distinguishes object algebras
from the traditional visitor pattern is the lack of a composite
interface with an \lstinline{accept} method, which
is both a blessing and a curse.  On
the one hand the trouble with composite interfaces with an
\lstinline{accept} method is that they make adding new variants to the
visitor pattern very hard. Although extensible versions of the visitor
pattern are possible, they usually require complex types using
advanced features of generics~\cite{togersen:2004,oliveira2012extensibility}. On the other hand, the lack of
such composite interfaces makes object algebras harder to use than
visitors. As illustrated in Section~\ref{subsec:OAs}, constructing expressions
with object algebras can only be done using a function parametrized by
an object algebra.

The remainder of the section shows that in \name there is no need to
have a dillema between extensibility using simple types and
usability: \emph{in \name it is possible to have extensible visitors with
simple types}! The key to achieve this is to have type inheritance
decoupled from subtyping, and allowing contravariant parameter type
refinement.

\subsubsection{The problem with extensible visitors}
\begin{comment}
is another design pattern that
facilitates extensibility. The gist of this pattern is to decouple an object
hierarchy from the behaviors of each object. In other words, objects no longer
contain operations but instead ``accept'' visitors to perform operations on
them. The visitor pattern tackles at the very problem that makes programming in
traditional OO styles hard to add a new operation, since operations are defined
inside object classes that represent data structures. As a result, using the
visitor pattern allows adding new operations to existing structures without
modifying code of structures, a style enjoyed by functional
programming.
\end{comment}
We illustrate the problem with extensible visitors using Scala.
The type for expressions is defined in Scala as:
\begin{lstlisting}{language=scala}
trait Exp {
  def accept[E](v: ExpAlg[E]): E
}
\end{lstlisting}
The trait \lstinline{Exp} has only one method
\lstinline$accept$, which takes an internal visitor (or object
algebra) as an argument.
Here the type \lstinline{ExpAlg[E]} is the Scala analogous of the
corresponding type defined in Section~\ref{subsec:OAs} in \name. In terms
of the visitor pattern, \lstinline{ExpAlg} defines the visit methods
for all variants.

\begin{comment}
The actual shape of the expressions
(i.e., variants) is determined by the type of the visitor, which we define in
another trait:
\begin{lstlisting}{language=scala}
trait ExpVisitor[E] {
  def lit(x: Int): E
  def add(e1: E, e2: E): E
}
\end{lstlisting}
\end{comment}

\paragraph{Adding a new variant.}
The difficulties arise when a new variant, such as subtraction is
added. To do so an extended visitor interface analogous to
\lstinline$SubExpAlg$ is needed. Moreover a corresponding composite
interface \lstinline$SubExp$ is needed as well:
%For example, \lstinline$SubExp$ and \lstinline$SubExpAlg$
%together represent the type of expressions supporting subtraction, in addition
%to literal and addition.
\begin{lstlisting}{language=scala}
trait SubExp extends Exp {
  def accept[E](v: SubExpAlg[E]): E
}
\end{lstlisting}
%We extend the original visitor and add the new case.
The body of \lstinline{Exp} and \lstinline{SubExp} are almost the same: they
both contain an \lstinline{accept} method that takes an object algebra
and returns a value of the type \lstinline{E}. The only difference in
\lstinline{SubExp} the object algebra
\lstinline{v} is of type \lstinline{SubExpAlg[E]}, which is a subtype of
\lstinline{ExpAlg[E]}.

\paragraph{Inheritance is not subtyping.}
Since \lstinline{v} appears in parameter position of
\lstinline{accept} and function parameters are naturally contravariant,
\lstinline{SubExp[E]} should be a \emph{supertype} (and not a subtype)
of \lstinline{Exp[E]}.  However, in Scala every extension induces a
subtype. In other words type inheritance and subtyping always go along
together.  To ensure type-soundness Scala (and other common OO
languages) forbids any kind of type-refinement on method parameter
types.  In other words method parameter types are invariant.
The consequence of this is that Scala is not capable of expressing
that \lstinline{SubExp[E]} is an extension and a supertype of
\lstinline{Exp}. Such kind of extension is an example where
``\emph{inheritance is not subtyping}``~\cite{cook1989inheritance}.

\begin{comment}
However,
such subtyping relation does not fit well in Scala because inheritance implies
subtyping in such languages \footnote{It is still possible to encode
  contravariant parameter types in Scala but doing so would require some
  technique.\bruno{what technique?}}. As \lstinline{SubExp[E]} extends
\lstinline{Exp[E]}, the former becomes a subtype of the latter. This suffers
from the ``Inheritance is Not Subtyping'' problem.
\end{comment}

\subsection{Extensible Visitors in \name}
Such limitation does not exist in \name. For example, we can define the similar
interfaces for \lstinline{Exp} and \lstinline{SubExp}:
\begin{comment}
  \begin{lstlisting}{language=F2J}
    type Exp    = {
      accept: forall A. ExpAlg[A] -> A
    };
    type SubExp = {
      accept: forall A. SubExpAlg[A] -> A
    };
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=7-12]{../src/Visitor.sf} % APPLY:linerange=V_EXP_SUBEXP
\name support contravariant parameter type refinement, which means that
\lstinline{SubExp} is a supertype of \lstinline{Exp}. Using these
types we first define two data constructors for simple expressions:
\begin{comment}
  \begin{lstlisting}{language=F2J}
    let lit (n: Int): Exp = {
      accept = /\E -> \(f: ExpAlg[E]) -> f.lit n
    };
    let add (e1: Exp) (e2: Exp): Exp = {
      accept = /\E -> \(f: ExpAlg[E]) ->
      f.add (e1.accept[E] f) (e2.accept[E] f)
    };
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=23-29]{../src/Visitor.sf} % APPLY:linerange=V_LITADD

\noindent Both \lstinline{lit} and \lstinline{add} build values of type
\lstinline{Exp} and use object algebras of type \lstinline{ExpAlg[E]}.
However, subtraction requires a value of type \lstinline{SubExp} to be created:
\begin{comment}
  \begin{lstlisting}{language=F2J}
    let sub (e1: SubExp) (e2: SubExp): SubExp ={
      accept = /\E -> \(f : SubExpAlg[E]) ->
      f.sub (e1.accept[E] f) (e2.accept[E] f)
    };
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=36-39]{../src/Visitor.sf} % APPLY:linerange=V_SUB

% One big benefit of using the visitor pattern is that programmers are able to
% define structures in a more natural way.

\paragraph{Usage.} With visitors constructing expressions is quite simple:

\begin{comment}
  \begin{lstlisting}{language=F2J}
    e2 = sub (lit 7) (lit 2)
  \end{lstlisting}
\end{comment}
\lstinputlisting[linerange=49-50]{../src/Visitor.sf} % APPLY:linerange=V_USAGE
The result is \lstinline$"7 - 2"$. Note that the programmer is able to pass \lstinline{lit 2}, which is of type \lstinline{Exp},
to \lstinline{sub}, which expects a \lstinline{SubExp}. The types are compatible
because because \lstinline$Exp$ is a \emph{subtype} of \lstinline$SubExp$. Code
reuse is achieved since we can use the constructors from \lstinline$Exp$ as the
constructor for \lstinline$SubExp$. In Scala, we would have to define two
literal constructors, one for \lstinline$Exp$ and another for
\lstinline$SubExp$.

Compared to object algebras, the addition of the composite structure
allows values to be created much more intuitively, without any
drawback! All the code developed with object algebras works right
away with visitors.

Finally note that in terms of typing, this solution does not require
any advanced use of generics. This is in sharp constrast with previous
proposals for extensible visitors in the literature.

% \subsection{Yanlin stuff}
% \bruno{This can be dropped.}

% This subsection presents yet another lightweight solution to the Expression
% Problem, inspired by the recent work by Wang. It has been shown that
% contravariant return types allows refinement of the types of extended
% expressions.

% First, we define the type of expressions that support evaluation and implement
% two constructors:
% \begin{lstlisting}
% type Exp = { eval: Int }
% let lit (n: Int) = { eval = n }
% let add (e1: Exp) (e2: Exp)
%   = { eval = e1.eval + e2.eval }
% \end{lstlisting}

% If we would like to add a new operation, say pretty printing, it is nothing more
% than refining the original \lstinline{Exp} interface by \emph{intersecting} the
% original type with the new \lstinline{print} interface using the \lstinline{&}
% primitive and \emph{merging} the original data constructors using the \lstinline{,,}
% primitive.
% \begin{lstlisting}
% type ExpExt = Exp & { print: String }
% let litExt (n: Int) = lit n ,, { print = n.toString() }
% let addExt (e1: ExpExt) (e2: ExpExt)
%   = add e1 e2 ,,
%     { print = e1.print.concat(" + ").concat(e2.print) }
% \end{lstlisting}

% Now we can construct expressions using the constructors defined above:
% \begin{lstlisting}
% let e1: ExpExt = addExt (litExt 2) (litExt 3)
% let e2: Exp = add (lit 2) (lit 4)
% \end{lstlisting}
% \lstinline{e1} is an expression capable of both evaluation and printing, while
% \lstinline{e2} supports evaluation only.

% We can also add a new variant to our expression:
% \begin{lstlisting}
% let sub (e1: Exp) (e2: Exp) = { eval = e1.eval - e2.eval }
% let subExt (e1: ExpExt) (e2: ExpExt)
%   = sub e1 e2 ,, { print = e1.print.concat(" - ").concat(e2.print) }
% \end{lstlisting}

% Finally we are able to manipulate our expressions with the power of both
% subtraction and pretty printing.
% \begin{lstlisting}
% (subExt e1 e1).print
% \end{lstlisting}

% \subsection{Mixins}

% \bruno{Still not convinced by this section. Change to the record-based example.}
% Mixins are useful programming technique wildly adopted in dynamic programming
% languages such as JavaScript and Ruby. But obviously it is the programmers'
% responsbility to make sure that the mixin does not try to access methods or
% fields that are not present in the base class.

% In Haskell, one is also able to write programs in mixin style using records.
% However, this approach has a serious drawback: since there is no subtyping in
% Haskell, it is not possible to refine the mixin by adding more fields to the
% records. This means that the type of the family of the mixins has to be
% determined upfront, which undermines extensibility.

% \name is able to overcome both of the problems: it allows composing mixins
% that (1) extends the base behavior, (2) while ensuring type safety.

% The figure defines a mini mixin library. The apostrophe in front of types
% denotes call-by-name arguments similar to the \lstinline{=>} notation in the
% Scala language.

% \begin{lstlisting}{language=F2J}
% type Mixin S = 'S -> 'S -> S;
% let zero S (super : 'S) (this : 'S) : S = super;
% let rec mixin S (f : Mixin S) : S
%   = let m = mixin S in f (\ (_ : Unit). m f) (\ (_ : Unit). m f);
% let extends S (f : Mixin S) (g : Mixin S) : Mixin S
%   = \ (super : 'S). \ (this  : 'S). f (\ (d : Unit). g super this) this;
% \end{lstlisting}

% We define a factorial function in mixin style and make a \lstinline{noisy} mixin
% that prints ``Hello'' and delegates to its superclass. Then the two functions
% are composed using the \lstinline{mixin} and \lstinline{extends} combinators.
% The result is the \lstinline{noisyFact} function that prints ``Hello'' every
% time it is called and computes factorial.
% \begin{lstlisting}{language=F2J}
% let fact (super : 'Int -> Int) (this : 'Int -> Int) : Int -> Int
%   = \ (n : Int). if n == 0 then 1 else n * this (n - 1)
% let noisy (super : 'Int -> Int) (this : 'Int -> Int) : Int -> Int
%   = \ (n : Int). { println("Hello"); super n }
% let noisyFact = mixin (Int -> Int) (extends (Int -> Int) foolish fact)
% noisy 5
% \end{lstlisting}

% \subsection{Composing Mixins and Object Algebras}
