\section{Applications}
\label{subsec:records}

\name is illustrated with two applications. The
first application shows how to mimic some of TypeScript's examples of 
dynamic mixins in \name. The second application shows how \name
enables a powerful form of polymorphic extensible records.

\subsection{Dynamic Mixins}\label{subsec:mixins}
%\subsection{Intersections in nominal type-systems}
%Recently several languages have embraced intersection types, such as Scala, Ceylon or TypeScript.
%In Scala, it is possible to define a merge-like function:
TypeScript is a language that adds static type checking to JavaScript.
Amongst numerous static typing constructs, \emph{TypeScript supports a form of intersection types, 
without a merge operator}. 
However, it is possible to define a function \lstinline{extend}
that mimics the merge operator:

\begin{lstlisting}
function extend<T, U>(first: T, second: U): T & U {
  let result = <T & U>{};
  for (let id in first) {
    (<any>result)[id] = (<any>first)[id];
  }
  for (let id in second) {
    if (!result.hasOwnProperty(id)) {
      (<any>result)[id] = (<any>second)[id];
    }
  }
  return result;
}

class Person { constructor(public name : string, public male : boolean) { } }
interface Loggable { log() : void; }
class ConsoleLogger implements Loggable { log() {...} }
var jim = extend(new Person("Jim",true), new ConsoleLogger());
var n = jim.name;
jim.log(); 
\end{lstlisting}

In this example, taken from TypeScript's documentation\footnote{We have added the field \lstinline$male$ to the class
\lstinline$Person$.}, 
an \lstinline$extend$ function is defined 
for mixin composition. Two classes \lstinline$Person$ and
\lstinline$ConsoleLogger$ are also defined.
Two instances of those classes are then composed in a variable \lstinline$jim$ with the type of 
the intersection of both using \lstinline{extend}. It is type-safe to access both the properties
from \lstinline$Person$ and \lstinline$ConsoleLogger$ in the object \lstinline$jim$.

TypeScript's implementation of \lstinline{extend} relies on a biased choice.
The function starts by creating a variable \lstinline$result$ with the type of the intersection. 
It then iterates through \lstinline$first$'s properties and copies them to 
\lstinline$result$.
Next, it iterates through \lstinline$second$'s properties but it only copies the properties
that \lstinline$result$ does not possess yet (i.e. the ones present in \lstinline$first$).
This means that the implementation is left-biased, as the properties of left type of the intersection
are chosen in favor of the ones in the right.
However, in TypeScript this may be a cause of severe problems since that, at the time of writing,
intersections at type-level are right-biased!
For example, the following code is well-typed:

\begin{lstlisting}[mathescape=true]
class Dog { constructor(public name : string, public male : string) { } }
var fool : Dog & Person = extend(new Dog("Pluto","yes"),new Person("Arnold",true));
boolean b = fool.male;     /* $\text{{\bf Undetected type-error here!}}$ */
\end{lstlisting}
There are a few problems here. Firstly both \lstinline{Dog} and
\lstinline{Person} contain a \lstinline{name} field, and the use of
\lstinline{extend} will favour the \lstinline{name} field in the first object. This
could be surprising for someone unfamiliar with the semantics of
\lstinline{extend} and, more importantly, it could easily allow
unintended name clashes to go undetected. 
Secondly, note how \lstinline$fool.male$ is statically
bound to a variable of type \lstinline$boolean$ but, at run-time, it
will contain a value of type \lstinline$String$! Thus the example
shows some run-time type errors can still occur when using \lstinline{extend}.

Other problematic issues regarding the semantics of intersection types can include the order of 
the types in an intersection, or even intersections including repeated types.
This motivates the need to define a clear meaning for the practical application of 
intersection types.

\paragraph{Dynamic mixins in \name}
In \name, the merge operator is built-in in \name. Thus \lstinline{extend} is simply defined as
follows:

\begin{lstlisting}
let extend T (U * T) (first : T, second : U) : T & U = first ,, second in 
\end{lstlisting}

\noindent The disjointness constraint on \lstinline{U} ensures that no
conflicts (such as duplicated fields of the same type) exists when merging the two objects. In practice this approach
is quite similar to trait-based OO approaches~\cite{scharli2003traits}. If conflicts
exist when two objects are composed, then they have to be resolved
manually (by dropping fields from some object, for example). Moreover if no existing
implementation can be directly reused, a new one must be provided via
record extension, analogously to standard method overriding in OO
languages.

For the previous TypeScript examples, assuming a straightforward
translation from objects to (polymorphic) records, then the
composition of \lstinline{person} and \lstinline{consoleLogger} is
well-typed in \name:

\begin{lstlisting}[mathescape=true]
type Person = {name : String} & {male : Bool};
type Loggable = {log : $\top \to \top$};

let person (n : String) (s : Bool) : Person = {name = n} ,, {male = s} in
let consoleLogger : Loggable = {log = ...} in
let jim = extend Person Loggable (person "Jim" true)  consoleLogger in
let n = jim.name in
jim.log $\top$
\end{lstlisting}

However, the intersection \lstinline$Dog & Person$ is not accepted.
This is due to both types sharing a field with the same name (\lstinline$name$) 
and the same type ($String$).
Note that the name clash between \lstinline$male$ fields (which have
different types) does not impose any 
problem in this example: \name allows and keeps duplicated fields
whose types are disjoint. This feature of \name is further illustrated next.

\subsection{Extensible Records}
\name can encode polymorphic extensible records. Describing and
implementing records within programming languages is certainly not
novel and has been extensively studied in the past, including
systems with row types \cite{wand1987complete,wand1989type}; 
predicates \cite{harper1990extensible,harper1991record,gaster1996polymorphic}; 
flags \cite{remy1993type}; conditional constraints \cite{pottier2003constraint};
cases \cite{blume2006extensible}; amongst others. However, while 
most systems have non-trivial built-in constructs to model various aspects of
records, \name specializes the more general notion of intersection
types to encode complex records.

\paragraph{Records and record operations in \name}
Systems with records usually rely on 3 basic operations: selection,
restriction and extension/concatenation.  Selection and concatenation
(via the merge operator) are built-in in the semantics of
\name. Merges in \name can be viewed as a generalization of record
concatenation.  In \name, following well-known encodings of
multi-field records in
systems with intersection types and a merge
operator~\cite{reynolds1997design,reynolds1991coherence}, there are only three rather simple constructs
for records: 1) single field record types; 2) single field records; 3)
field accessors. Multi-field records in \name are encoded with
intersections and merges of single field records. An example is
already illustrated in Section~\ref{subsec:mixins}.  The record type
\lstinline$Person$ is the intersection of two single field record
types. The record \lstinline$person "Jim" true$ is built with a merge
of two single field records. Finally, \lstinline{jim.name} and
\lstinline{jim.log} illustrates the use of field accessors.  Note how,
through the use of subtyping, accessors will accept any intersection
type that contains the single record with the corresponding field.
This resembles systems with record subtyping
\cite{cardelli1990operations,pierce1994simple}.


\begin{comment}
\paragraph{Record terms and types}
Our system directly encodes a term for the single record construct $\recordCon l e$, where $l$ is some
label and $e$ is some other term.
This term comes with its associated type, denoted as $\recordType l T$, where $T$ is a type that is attributed 
to $e$. 

\paragraph{Selection}
The select operator is directly embedded in our language. 
It follows the usual syntax of $e.l$, where $e$ is an expression of type $\recordType l \alpha$ and 
$l$ is a label.
A polymorphic function which extracts any record that include the label $l$ of type
$\alpha$ could be written as:
%\begin{align*}
%&\text{select} :: \fordis \alpha \top {\recordType l \alpha \rightarrow \alpha} \\
%&\text{select} \, = \blamdis \alpha \top {\lam x {x.l}}
%\end{align*}

\begin{lstlisting}[mathescape=true]
let select A (r : {l : A}) : A = r.l in $\ldots$ 
\end{lstlisting}

Note how, through the use of subtyping, this function will accept any intersection type that 
contains the single record $\recordType l A$.
This may resemble other systems with subtyping \cite{cardelli1990operations,pierce1994simple}, 
although it is slightly more general -- the type $A$ is not restricted only  
restricted to record types. 
\end{comment}

\paragraph{Restriction via subtyping}
In contrast to most record systems, restriction is not directly
embedded in \name.
Instead, \name uses subtyping for restriction: 
%\begin{align*}
%&\text{remove} :: \fordis \alpha \top {\fordis r {\recordType l \alpha} 
%{({\recordType l \alpha \inter r) \rightarrow r}}} \\ 
%&\text{remove} \, = \blamdis \alpha \top {\blamdis r {\recordType l \alpha} {\lam x {x}}} 
%\end{align*}
%
\begin{lstlisting}[mathescape=true]
let remove (x : {age : Int} & {name : String}) : {name : String} = x in $\ldots$ 
\end{lstlisting}

\noindent The function \lstinline{remove} drops the field
\lstinline{age} from the record \lstinline{x}. 

\paragraph{Polymorphic extensible records} Records in \name can have
polymorphic fields, and disjointness enables encoding various
operations expressible in systems with polymorphic records. For example, 
the following variant of \lstinline{remove}

\begin{lstlisting}[mathescape=true]
let remove A (B * {l : A}) (x : { l : A } & B) : B = x in $\ldots$ 
\end{lstlisting}

\noindent takes a value \lstinline{x} which contains a record of type
\lstinline{l : A}, as well as some extra information of type
\lstinline{B}. The disjointness constraint on \lstinline{B} ensures
that values of type \lstinline{B} do not contain a record with type
\lstinline{l : A}.  This example shows that one
can use disjoint quantification to express negative field information,
which is very close to the system described by Harper and
Pierce~\cite{harper1991record}.  Note, however, that \name requires
explicitly stating the type of field in the constraint, whereas
systems with a \emph{lacks} (field) predicate only require the name of the field.
The generality of disjoint intersection types, which allows one to
encode record types, is exactly what forces us to add this extra type
in the constraint.  However, there is a slight gain with \name's
approach: $remove$ allows \lstinline{B} to contain fields with label $l$, as long as
the field types are \emph{disjoint} to $\tyint$. Such fine-grained
constraint is not possible to express only with a lacks predicate.

\begin{comment}
\paragraph{Extension/Concatenation}
The most usual operators for combining records are extension and concatenation.
Even though that in some systems, the latter is defined in terms of the former, languages that
opt to include concatenation usually rely on specific semantics for it. 
Our system is suitable for encoding both of these operations, but we argue that concatenation is
the natural primitive operator, due to the resemblance with our merge
operator.
Indeed, \cite{harper1991record} also define a \emph{merge} operator, which is quite similar to
our \emph{merge} for intersection types, except it involves only record types.
For instance, in their system, the following function concatenates a single record with field $l$ 
of type $\tyint$ with another record that lacks this field \footnote{The syntax is slightly modified
with respect to the original notation}:
%\begin{align*}
%&\text{addL}_1 :: \forall \alpha\#l. \alpha \rightarrow (\alpha || \recordType l \tyint) \\ 
%&\text{addL}_1 \, = ...
%\end{align*}
\begin{lstlisting}[mathescape=true]
let $\text{addL}_1$ (A # l) (x : A) : (a || { l : Int }) = $\ldots$ in $\ldots$ 
\end{lstlisting}

The reader might notice the resemblance with our system:
%\begin{align*}
%&\text{addL}_2 :: \fordis \alpha {\recordType l \tyint} {
%\alpha \rightarrow (\alpha \inter \recordType l \tyint)} \\
%&\text{addL}_2 \, = ...
%\end{align*}
\begin{lstlisting}[mathescape=true]
let $\text{addL}_2$ (A * { l : Int }) (x : A) : (A & { l : Int }) = $\ldots$ in $\ldots$ 
\end{lstlisting}

This shows that one can use disjoint quantification to express negative field information, which
is very close to the system described in \cite{harper1991record}.  
Note how one has to explicitly state the type of the constraint in $\text{addL}_2$, whereas 
$\text{addL}_1$ does not require this.
The generality of disjoint intersection types, that allows one to encode record types, is exactly 
what forces us to add this extra type in the constraint.
However, there is a slight gain with this approach: $\text{addL}_2$ accepts more types than
$\text{addL}_1$.
Namely, all (intersection) types that contain label $l$, with a field type \emph{disjoint} to
$\tyint$.
Had one meant to forbid records with \emph{any} $l$ fields, then one could write:

%\begin{align*}
%&\text{addL}_3 :: \fordis \beta \top {\fordis \alpha {\recordType l \beta} {
%\alpha \rightarrow (\alpha \inter \recordType l \beta)}} \\
%&\text{addL}_3 \, = ...
%\end{align*}
\begin{lstlisting}[mathescape=true]
let $\text{addL}_3$ (A * { l : $\bot$ }) (x : A) : A & { l : $\bot$ } = $\ldots$ in $\ldots$ 
\end{lstlisting}
Unfortunately, our system does not support the $\bot$ type, so this program would not be accepted.
We will get into the reasons for this in further detail, in Section~\ref{sec:conclusion}.
\end{comment}


\paragraph{Expressibility}
As noted by Leijen~\cite{leijen2005extensible}, systems can typically be
categorized into two distinct groups in what concerns extension:
strict; and free. The former does not allow field overriding when
extending a record (i.e. one can only extend a record with a field
that is not present in it); while the latter does account for field
overriding.  Our system can be seen as hybrid of these two kinds of
systems.  

With \emph{lightweight extensible records}~\cite{jones99lightweight} -- a system with strict extension -- an example of a function that uses 
record types is the following:
%\begin{align*}
%&\text{avg}_1 :: (r \backslash y, r \backslash x) \Rightarrow \{r | x :: \tyint, y :: \tyint\} \rightarrow \tyint \\
%&\text{avg}_1 \, r = (r.x + r.y) / 2
%\end{align*}
\begin{lstlisting}[mathescape=true]
let $\text{avg}_1$ (R\y, R\y) => (r : {R | x:Int, y:Int}) = (r.x+r.y)/2
\end{lstlisting}
The type signature says that any record $r$, containing fields $x$
and $y$ and some more information $R$ (which lacks both fields $x$ and
$y$), can be accepted returning an integer.  Note how the bounded
polymorphism is essential to ensure that $R$ does not contain $x$ nor
$y$.

On the other hand, in Leijen's~\cite{leijen2005extensible} system with free extension 
the more general program would be accepted:
%\begin{align*}
%&\text{avg}_2 :: \forall \; r \;x \; y, \{ x :: \tyint, y :: \tyint | r \} \rightarrow \tyint \\
%&\text{avg}_2 \, r = (r.x + r.y) / 2
%\end{align*}
\begin{lstlisting}[mathescape=true]
let $\text{avg}_2$ R (r : {x:Int, y:Int | R}) = (r.x+r.y)/2 
\end{lstlisting}
In this case, if $R$ contains either field $x$ or field $y$, they would be shadowed by the labels 
present in the type signature.
In other words, in a record with multiple $x$ fields, the most recent (i.e. left-most) is used 
in any function which accesses $x$.

In \name the following program can written instead:
%\footnote{We do not support exactly this function definition style; 
%however the type signature and expression (modulo infix operators) are exactly as one would write them
%in \name}:
%\begin{align*}
%&\text{avg}_3 :: \fordis r 
%{\recordType x \tyint \inter \recordType y \tyint} 
%{\recordType x \tyint \inter \recordType y \tyint \inter r \rightarrow \tyint} \\
%&\text{avg}_3 = 
%{\blamdis t {\recordType x \tyint \inter \recordType y \tyint} {\lam r {(r.x + r.y) / 2}}}
%\end{align*}
\begin{lstlisting}[mathescape=true]
let $\text{avg}_3$ (R*{x:Int}&{y:Int}) (r : {x:Int}&{y:Int}&R) = (r.x+r.y)/2 
\end{lstlisting}
Since \name accepts duplicated fields as long as the types of the
overlapping fields are disjoint, more inputs are accepted by this
function than in the first system. However, since Leijen's system
accepts duplicated fields even when types are overlapping, $\text{avg}_3$
accepts less types than $\text{avg}_2$. 
Another major difference between \name and the two other mentioned systems, is the ability to 
combine records with arbitrary types.
Our system does not account for well-formedness of record types as the other two systems do 
(i.e. using a special \emph{row} kind), since our encoding of records piggybacks on the more
general notion of disjoint intersection types. 

\begin{comment}
Finally, it is also worth noting that systems using subtyping may suffer from the so-called 
\emph{update} problem~\cite{cardelli1990operations}.
\name does not suffer from this problem as it does not support refinement types. 
We may illustrate this by defining a suitable update function, in a similar fashion 
to~\cite{leijen2005extensible}:
%\begin{align*}
%&\text{update} :: \fordis \alpha \top {\fordis r {\recordType l \alpha} 
%{{\recordType l \alpha \inter r \rightarrow \beta \rightarrow \recordType l \beta \inter r}}} \\ 
%&\text{update} \, = \blamdis \alpha \top {\blamdis r {\recordType l \alpha} {\lam x {\lam v {
%\recordCon l v \mergeop (\text{remove } \alpha \; r \; x)}}}} 
%\end{align*}
\begin{lstlisting}[mathescape=true]
let $\text{update}$ A B (R * { l : A }) (r : { l : A } & R) (v : B) : { l : B } & R = { l = v } ,, (remove a r x) in $\ldots$ 
\end{lstlisting}
\edn{comment}

\begin{comment}
\section{Application: Extensibility}
\label{subsec:OAs}

Various solutions to the Expression
Problem~\cite{wadler1998expression} in the
literature~\cite{Swierstra:2008,finally-tagless,oliveira09modular,oliveira2012extensibility,DelawareOS13}
are closely related to type-theoretic encodings of datatypes. Indeed, variants
of the same idea keep appearing in different programming languages,
because the encoding of the idea needs to exploit the particular
features of the programming language (or theorem prover).
Unfortunately language-specific constructs obscure the key ideas
behind those solutions.

In this section we show a solution to the
Expression Problem that intends to capture the key ideas of various
solutions in the literature. Moreover, it is shown how \emph{all the
  features} of \namedis (intersection types, the merge operator,
parametric polymorphism and disjoint quantification) are needed to
properly encode one important combinator~\cite{oliveira2013feature} used to compose
multiple operations over datatypes.

%The combination of parametric polymorphism, intersection types and the
%merge operator is useful to encode datatypes with subtyping and
%support extensibility in \namedis. The combination of
%polymorphism and the merge operator is essential to allow the
%composition of multiple operations over datatypes.
%In the following
%we present a step-by-step solution to the Expression Problem in \namedis,
%and illustrate how to combine multiple operations.

%Oliveira and Cook~\cite{oliveira2012extensibility} proposed a design pattern that can solve the
%Expression Problem in languages like Java. An advantage of the pattern
%over previous solutions is that it is relatively lightweight in terms
%of type system features. In a latter paper, Oliveira et al.~\cite{oliveira2013feature}
%noted some limitations of the original design pattern and proposed
%some new techniques that generalized the original pattern, allowing it
%to express programs in a Feature-Oriented Programming~\cite{Prehofer97} style.
%Key to these techniques was the ability to dynamically compose object
%algebras.
%
%Unfortunatelly, dynamic composition of object algebras is
%non-trivial. At the type-level it is possible to express the resulting
%type of the composition using intersection types. Thus, it is still
%possible to solve that part problem nicely in a language like Scala (which
%has basic support for intersection types). However, the dynamic
%composition itself cannot be easily encoded in Scala. The fundamental
%issue is that Scala lacks a \lstinline{merge} operator (see the
%discussion in Section~\ref{subsec:interScala}). Although both Oliveira et al.~\cite{oliveira2013feature} and
%Rendell et al.~\cite{rendel14attributes} have shown that such a \lstinline{merge} operator can
%be encoded in Scala, the encoding fundamentally relies in low-level
%programming techniques such as dynamic proxies, reflection or
%meta-programming.
%
%Because \namedis supports a \lstinline{merge} operator natively, dynamic
%object algebra composition becomes easy to encode. The remainder of
%this section shows how object algebras and object algebra composition
%can be encoded in \namedis. We will illustrate this point
%step-by-step by solving the Expression Problem.
%%Prior knowledge of object algebras is not assumed.

% can be cumbersome and
% language support for intersection types would solve that problem.
% Our type system is just a simple extension of System $ F $; yet surprisingly, it
% is able to solve the limitations of using object algebras in languages such as
% Java and Scala.

\subsection{Church Encoded Arithmetic Expressions}
In the Expression Problem, the idea is to start with a very simple
system modelling arithmetic expressions and evaluation.
The standard typed Church encoding ~\cite{BoehmBerarducci} for arithmetic expressions,
denoted as the type \lstinline{CExp}, is:

\begin{lstlisting}
type CExp = (*$ \forall $*)E. (Int (*$ \to $*) E) (*$ \to $*) (E (*$ \to $*) E (*$ \to $*) E) (*$ \to $*) E
\end{lstlisting}

\noindent However, as done in various solutions to extensibility, it is better to break
down the type of the Church encoding into two parts:

\begin{lstlisting}
type ExpAlg[E] = {
  lit: Int (*$ \to $*) E,
  add: E (*$ \to $*) E (*$ \to $*) E
} in (*$ \ldots $*)
\end{lstlisting}

\noindent The first part, captured by the type \lstinline{ExpAlg[E]}
is constitutes the so-called algebra of the datatype. For additional
clarity of presentation, records (supported in the implementation of \namedis)
are used to capture the two components of the algebra.
The first component abstracts over the type of the
constructor for literal expressions ($\tyint \to E$). The second
component abstracts over the type of addition expressions
($E \to E \to E$).

The second part, which is the actual type of the Church encoding, is:

\begin{lstlisting}
type Exp = { accept: (*$ \forall $*)E. ExpAlg[E] (*$ \to $*) E } in (*$ \ldots $*)
\end{lstlisting}

\noindent It should be clear that, modulo some re-factoring, and the
use of records, the type \lstinline{Exp} and \lstinline{CExp}
are equivalent.

\paragraph{Data constructors.}
Using \lstinline{Exp} the two data constructors are defined as follows:

\begin{lstlisting}
let lit (n: Int): Exp = {
  accept = (*$ \Lambda $*)E (*$ \to $*) fun (f: ExpAlg[E]) (*$ \to $*) f.lit n
} in
let add (e1: Exp) (e2: Exp): Exp = {
  accept = (*$ \Lambda $*)E (*$ \to $*) fun (f: ExpAlg[E]) (*$ \to $*)
    f.add (e1.accept[E] f) (e2.accept[E] f)
} in
(*$ \ldots $*)
\end{lstlisting}

Note that the notation $\Lambda E$ in the definition of the
\lstinline{accept} fields is a type abstraction: it
introduces a type variable in the environment. The definition of the
constructors themselves follows the usual Church encodings.

Simple expressions, can be built using the data constructors:

\begin{lstlisting}
let five : Exp = add (lit 3) (lit 2) in (*$ \ldots $*)
\end{lstlisting}

\paragraph{Operations.} Defining operations over expressions requires
implementing \lstinline{ExpAlg[E]}. For example, an interesting
operation over expressions is evaluation. The first step is to define
the evaluation operation is to chose how to instantiate the type
parameter \lstinline{E} in \lstinline{ExpAlg[E]} with a suitable
concrete type for evaluation. One such suitable type is:

\begin{lstlisting}
type IEval = { eval: Int } in (*$ \ldots $*)
\end{lstlisting}

\noindent Using \lstinline{IEval}, a record \lstinline{evalAlg}
implementing \lstinline{ExpAlg} is defined as follows:

\begin{lstlisting}
let evalAlg: ExpAlg[IEval] = {
  lit = fun (x: Int) (*$ \to $*) { eval = x },
  add = fun (x: IEval) (y: IEval) (*$ \to $*) {
    eval = x.eval + y.eval
  }
} in (*$ \ldots $*)
\end{lstlisting}

In this record, the two operations
\lstinline{lit} and \lstinline{add} return a record with type
\lstinline{IEval}. The definition of \lstinline{eval} for
\lstinline{lit} and \lstinline{add} is straightforward.


%The type \lstinline$ExpAlg[IEval]$ is the type of object algebras
%supporting evaluation.
%However, the one interesting point
%of object algebras is that other operations can be supported as
%well.

Using \lstinline{evalAlg}, the expression \lstinline{five} can
be evaluated as follows:

\begin{lstlisting}
(five.accept[IEval] evalAlg).eval
\end{lstlisting}

\subsection{Extensibility and Subtyping} Of course, in the Expression
Problem the goal is to achieve extensibility in two dimensions: constructors and operations.
Moreover, in the presence of subtyping it is also interesting to see how the extended datatypes
relate to the original datatypes. We discuss the two topics next.

\paragraph{New constructors.} Here is the code needed to add a new subtraction constructor:
%Arithmetic expressions with subtyping are defined using the type \lstinline{SubExp}:

\begin{lstlisting}
type SubExpAlg[E] = ExpAlg[E] & { sub: E (*$ \to $*) E (*$ \to $*) E } in
type SubExp = { accept: (*$ \forall $*)A. SubExpAlg[A] (*$ \to $*) A } in
let sub (e1: SubExp) (e2: SubExp): SubExp = {
  accept = (*$ \Lambda $*)E (*$ \to $*) fun (f : SubExpAlg[E]) (*$ \to $*)
    f.sub (e1.accept[E] f) (e2.accept[E] f)
} in
(*$ \ldots $*)
\end{lstlisting}

\noindent Firstly \lstinline{SubExpAlg} defines an extended algebra
that contains the constructors of \lstinline{ExpAlg} plus the new
subtraction constructor. Intersection types are used to do the type
composition. Secondly, a new type of expressions with subtraction
(\lstinline{SubExp}) is needed. For \lstinline{SubExp} it is important
that the \lstinline{accept} field now takes an algebra of type
\lstinline{SubExpAlg} as argument. This is necessary to define the
constructor for subtraction (\lstinline{sub}), which requires the
algebra to have the field \lstinline{sub}.


\paragraph{Extending existing operations.} In order to use evaluation
with the new type of expressions, it is necessary to also extend
evaluation. Importantly, extension is achieved using the merge operator:

\begin{lstlisting}
let subEvalAlg = evalAlg ,, {
  sub = fun (x: IEval) (y: IEval) (*$ \to $*) {
    eval = x.eval - y.eval
  }
} in (*$ \ldots $*)
\end{lstlisting}

\noindent In the code, the merge operator takes \lstinline{evalAlg}
and a new record with the implementation of evaluation for
subtraction, to define the implementation for arithmetic expressions
with subtraction.

\paragraph{Subtyping.} In the presence of subtyping, there are
interesting subtyping relations between datatypes and their
extensions~\cite{oliveira09modular}.  Such subtyping relations are usually not
discussed in theoretical treatments of Church encodings. This is
probably partly due to most work on typed Church encodings being
done in calculi without subtyping.

The interesting aspect about subtyping in typed Church encodings is
that subtyping follows the opposite direction of the extension.  In
other words subtyping is contravariant with respect to the
extension. Such contravariance is explained by the type of the
\lstinline{accept} field, which is a function where the argument type
is refined in the extensions. Thus, due to the contravariance of
subtyping on functions, the extension becomes a supertype of the
original datatype.

In the particular case of expressions \lstinline{Exp} (the original and smaller
datatype) is a subtype of \lstinline{SubExp} (the larger and extended
datatype). Because of this subtyping relation, writing the following
expression is valid in \namedis:

\begin{lstlisting}
let three : SubExp = sub five (lit 2)
\end{lstlisting}

\noindent Note the \lstinline{three} is of type \lstinline{SubExp},
but the first argument (\lstinline{five}) to the constructor
\lstinline{sub} is of type \lstinline{Exp}. This can only type-check
if \lstinline{Exp} is indeed a subtype of \lstinline{SubExp}.





\paragraph{New operations.}
The second type of extension is adding a new operation, such as pretty printing.
Similarly to evaluation, the interface of the pretty printing feature
is modeled as:
%\begin{comment}
%  \begin{lstlisting}
%    type IPrint = { print : String } in (*$ \ldots $*)
%  \end{lstlisting}
%\end{comment}
\begin{lstlisting}
type IPrint = { print: String } in (*$ \ldots $*)
\end{lstlisting}
The implementation of pretty printing for expressions that support literals,
addition, and subtraction is:

\begin{lstlisting}
let printAlg: SubExpAlg[IPrint] = {
  lit = fun (x: Int) (*$ \to $*) {
    print = x.toString()
  },
  add = fun (x: IPrint) (y: IPrint) (*$ \to $*) {
    print = x.print ++ " + " ++ y.print
  },
  sub = fun (x: IPrint) (y: IPrint) (*$ \to $*) {
    print = x.print ++ " - " ++ y.print
  }
} in (*$ \ldots $*)
\end{lstlisting}

\noindent The definition of \lstinline{printAlg} is unremarkable.
With \lstinline{printAlg} we can pretty print the expression represented
by \lstinline{three}:

\begin{lstlisting}
(three.accept[IPrint] printAlg).print
\end{lstlisting}

%\begin{comment}
%The result is \lstinline$"7 - 2"$. Note that the programmer is able to pass \lstinline{lit 2}, which is of type \lstinline{Exp},
%to \lstinline{sub}, which expects a \lstinline{SubExp}. The types are compatible
%because because \lstinline$Exp$ is a \emph{subtype} of \lstinline$SubExp$. Code
%reuse is achieved since we can use the constructors from \lstinline$Exp$ as the
%constructor for \lstinline$SubExp$. In Scala, we would have to define two
%literal constructors, one for \lstinline$Exp$ and another for
%\lstinline$SubExp$.
%\end{comment}
%
\subsection{Composition of Algebras}
The final example shows a non-trivial combinator for algebras that
allows multiple algebras to be combined into one. A version of this
combinator has been encoded in Scala before using intersection types
(which Scala supports) and an encoding of the merge
operator~\cite{oliveira2013feature,rendel14attributes}.
Unfortunately, the Scala encoding of the merge operator is quite complex as
it relies on low-level type-unsafe programming features such as
dynamic proxies, reflection or other meta-programming techniques.
In \namedis there is no need for such hacky encoding, as the
merge operator is natively supported. Therefore the combinator for
composing algebras is implemented much more elegantly.
The combinator is defined by the \lstinline$combine$ function, which takes two object algebras to create
a combined algebra. It does so by constructing a new algebra
where each field is a function that delegates the input to the two
algebra parameters.

\begin{lstlisting}
let combine A (B * A) (f: ExpAlg[A]) (g: ExpAlg[B]) :
  ExpAlg[A & B] = {
    lit = fun (x: Int) (*$ \to $*) f.lit x ,, g.lit x,
    add = fun (x: A & B) (y: A & B) (*$ \to $*)
      f.add x y ,, g.add x y
}
\end{lstlisting}

Note how \lstinline{combine} requires all the interesting
features of \namedis. Parametric polymorphism is needed because
\lstinline{combine} must compose algebras with arbitrary type
parameters. Intersection types are needed because the resulting
algebra will create values with an intersection type composing
the two type parameters of the two input algebras. The merge operator
is needed to compose the results of each algebra together. Finally,
a disjointness constraint is needed to ensure that the two input
algebras build values of disjoint types (otherwise ambiguity could
arize).

With \lstinline{combine} printing and evaluation of expressions with
subtraction is done as follows:

\begin{lstlisting}
let newAlg: ExpAlg[IEval&IPrint] =
    combine[IEval,IPrint] evalAlg printAlg in
let o = five.accept[IEval&IPrint] newAlg in
o.print ++ " = " ++ o.eval.toString()
\end{lstlisting}

Note that \lstinline$o$ is a value that supports both
evaluation and printing. The final expression uses \lstinline{o}
for doing both printing and evaluation.
\end{comment}
