\section{Disjointness and Coherence}
\label{sec:disjoint}

Although the system shown in the Section~\ref{sec:fi} is type-safe, it is not
coherent. This section shows how to modify \name so that it guarantees coherence
as well as type soundness. The result is a calculus named \namedis. The keys
aspects are the notion of disjoint intersections, and disjoint quantification
for polymorphic types.

%From the theoretical point-of-view, the end goal of this section is to show that the resulting system has
%a coherent (or unique) elaboration semantics:
%\begin{restatable}[Unique elaboration]{theorem}{uniqueelaboration}
%  \label{theorem:unique-elaboration}
%
%  If $\jtype \Gamma e {A_1} \yields {E_1}$ and $\jtype \Gamma e {A_2} \yields
%  {E_2}$, then $E_1 \equiv E_2$. (``$\equiv$'' means syntactical equality, up to
%  $\alpha$-equality.)
%
%\end{restatable}
%
%\noindent In other words, given a source term $e$, elaboration always produces
%the same target term $E$. The most important hurdle we need to overcome is that
%if $A \inter B \subtype C$, then either $A$ or $B$ contributes to that subtyping
%relation, resulting in two possible coercions.

\subsection{Disjointness} Throughout the paper we already presented an intuitive
definition for disjoiness. Here such definition is made a bit more precise, and
well-suited to \namedis.

\begin{definition}[Disjoint types]

  Given a context $\Gamma$, two types $A$ and $B$ are said to be disjoint
  (written $\jdis \Gamma A B$) if they do not share a common supertype. That is,
  there does not exist a type $C$ such that $A \subtype C$ and that $B \subtype
  C$. Note that we assume that all free type variables in $A$,
    $B$ and $C$ are bound in $\Gamma$.

  \[\jdis \Gamma A B \equiv \not \exists C.~A \subtype C \wedge B \subtype C\]

\end{definition}

To see this definition in action, $\tyint$ and $\tychar$ are disjoint,
because there is
no type that is a supertype of the both. On the other hand, $\tyint$ is not
disjoint with itself, because $\tyint \subtype \tyint$. This implies that
disjointness is not reflexive as subtyping is. Two types with different shapes
are always disjoint, unless one of them is an intersection type. For example, a function
type and a universally quantified type must be disjoint. But a function type and an intersection
type may not be. Consider:
\[ \tyint \to \tyint \quad \text{and} \quad (\tyint \to \tyint) \inter (\tystring \to \tystring) \]
Those two types are not disjoint since $\tyint \to \tyint$ is their common supertype.

\subsection{Syntax}

\begin{figure}
  \[
    \begin{array}{l}
      \begin{array}{llrll}
        \text{Types}
        & A, B & \Coloneqq & \alpha                  & \\
        &      & \mid & \highlight {$\bot$}          & \\
        &      & \mid & A \to B                      & \\
        &      & \mid & \for {(\alpha \highlight {$\disjoint A$})} B  & \\
        &      & \mid & A \inter B                   & \\

        \\
        \text{Terms}
        & e & \Coloneqq & x                        & \\
        &   & \mid & \lam {(x \oftype A)} e          & \\
        &   & \mid & \app {e_1} {e_2}              & \\
        &   & \mid & \blam {(\alpha \highlight {$\disjoint A$})} e  & \\
        &   & \mid & \tapp e A                     & \\
        &   & \mid & e_1 \mergeop e_2              & \\

        \\
        \text{Contexts}
        & \Gamma & \Coloneqq & \cdot
                   \mid \Gamma, \alpha \highlight {$\disjoint A$}
                   \mid \Gamma, x \oftype A  & \\

        \text{Syntactic sugar} & \blam \alpha e & \equiv & \blamdis \alpha \bot e & \\
                               & \forall \alpha.~A & \equiv & \forall (\alpha * \bot)~.~A & \\
      \end{array}
    \end{array}
  \]

  \caption{Amendments of the rules.}
  \label{fig:fi-syntax-dis}
\end{figure}

\begin{figure}
  \framebox{$\im A = T$}

  \begin{align*}
    \im \bot                  &= () \\
    \im {\fordis \alpha A B}  &= \for \alpha \im B \\
  \end{align*}

  \framebox{$\im \Gamma = G$}

  \begin{align*}
    \im {\Gamma, \alpha \disjoint A} &= \im \Gamma, \alpha \\
  \end{align*}

  \caption{Additional type and context translation.}
  \label{fig:additional-type-and-context-translation}
\end{figure}

% \george{May also note on the scoping of type variables inside contexts.}

Figure~\ref{fig:fi-syntax-dis} shows the updated syntax with the
changes highlighted and Figure~\ref{fig:additional-type-and-context-translation}
shows type and context translations for the new constructs.
Note how similar the changes are to those needed
to extend System $F$ with bounded quantification. First, type
variables are now always associated with their disjointness
constraints (like $\alpha \disjoint A$) in types, terms, and
contexts. Second, the bottom type ($\bot$) is introduced so that
universal quantification becomes a special case of disjoint
quantification: $\blam \alpha e$ is really a syntactic sugar for
$\blamdis \alpha \bot e$. The underlying idea is that any type is
disjoint with the bottom type.  Note the analogy with bounded
quantification, where the top type is the trivial upper bound in
bounded quantification, while the bottom type is the trivial
disjointness constraint in disjoint quantification.

%Indeed, \bruno{unfinidhed sentence}\george{Mabe show a diagram here to contrast
%with bounded polymorphism.}

\subsection{Typing}
Figure~\ref{fig:fi-type-patch} shows modifications to Figure~\ref{fig:fi-type} in
order to support disjoint intersection types and disjoint
quantification. Only new rules or rules that different are shown.
Importantly, the disjointness judgement appears in the well-formedness rule for intersection
types and the typing rule for merges.

\begin{figure*}
  \begin{mathpar}
    \framebox{$\jatomic A$} \\

    \inferrule*
    {}
    {\jatomic \bot}

    \inferrule*
    {}
    {\jatomic {A \to B}}

    \inferrule*
    {}
    {\jatomic {\fordis \alpha B A}}
  \end{mathpar}

  \begin{mathpar}
    \formsub \\ \rulesubinterldis \and \rulesubinterrdis \and \rulesubforalldis
  \end{mathpar}

  \begin{mathpar}
    \formwf \\ \rulewfforalldis \and \rulewfinterdis
  \end{mathpar}

  \begin{mathpar}
    \formt \\ \ruletblamdis \and \rulettappdis \and \ruletmergedis
  \end{mathpar}

  \caption{Affected rules.}
  \label{fig:fi-type-patch}
\end{figure*}

\begin{figure*}
  \begin{mathpar}
    \framebox{$\jatomic A$} \\

    \inferrule*{}{\jatomic {A \to B}}

    \inferrule*{}{\jatomic \alpha}

    \inferrule*{}{\jatomic {\for \alpha A}}
  \end{mathpar}

  \begin{mathpar}
    \formsub \\ \rulesubint \and \rulesubvar \and \rulesubfun \and \rulesubforall
    \and \rulesubinter  \and \rulesubinterldis \and \rulesubinterrdis 
  \end{mathpar}

  \begin{mathpar}
    \formwf \\ \rulewfint \and \rulewfvar \and \rulewffun \and \rulewfforall \and \rulewfinter
  \end{mathpar}

%  \begin{mathpar}
%    \formt \\ \ruletvar \and \ruletlam \and \ruletapp \and
%    \ruletblam \and \rulettapp \and \ruletmergedis 
%  \end{mathpar}

  \begin{mathpar}
    \formbi \\ \bruletint \and \bruletvar \and \bruletapp \and
    \brulettapp \and \bruletmergedis \and \bruletann 
  \end{mathpar}

  \begin{mathpar}
    \formbc \\ \bruletlam \and  \bruletblam \and \bruletsub
  \end{mathpar}

  \caption{Rules for Naive system.}
  \label{fig:fi-type-naive}
\end{figure*}

\begin{figure*}
  \begin{mathpar}
    \framebox{$\jatomic A$} \\

    \inferrule*{}{\jatomic {\fordis \alpha B A}}
  \end{mathpar}

  \begin{mathpar}
    \formsub \\ \rulesubtop \and \rulesubforalldis \and 
    \rulesubinterlcoerce \and \rulesubinterrcoerce
  \end{mathpar}

  \begin{mathpar}
    \formwf \\ \rulewftop \and \rulewfvardis \and \rulewfforalldis 
  \end{mathpar}

  \begin{mathpar}
    \formbi \\ \brulettop \and \brulettappdis 
  \end{mathpar}

  \begin{mathpar}
    \formbc \\ \bruletblamdis 
  \end{mathpar}


  \caption{Changes for Extended systems.}
  \label{fig:fi-type-extended}
\end{figure*}

\begin{figure*}
  \begin{mathpar}
    \formsub \\ \rulesubforallext
  \end{mathpar}

  \caption{Rules for Extended system with improved ForAll.}
  \label{fig:fi-type-extended_forall}
\end{figure*}

\begin{figure*}[t]
  \begin{mathpar}
    \formtoplike \\ %\framebox{$\jatomic A$} \\

    \ruletopltop \and \ruletoplinter \and \ruletoplfun \and \ruletoplforall
  \end{mathpar}


  \begin{center}
    \framebox{$\andcoerce{A}_{C} = T$}
  \end{center}

  \[
  \andcoerce{A}_{C} = 
  \begin{cases} 
        \toplike{A} & \andcoerce{A} \\ 
        %A = \top & () \\
        %A = A_1 \to A_2 \; \wedge \; \toplike{A_2} & \lam x \andcoerce{A_2}_{C} \\
        \text{otherwise} & {C} 
  \end{cases}
  \]

  \begin{center}
    \framebox{$\andcoerce{A} = T$}
  \end{center}

  \[
  \andcoerce{A} = 
  \begin{cases} 
        A = \top & () \\
        A = A_1 \to A_2 & \lam x \andcoerce{A_2} \\
        A = A_1 \inter A_2 & \pair {\andcoerce{A_1}} {\andcoerce{A_2}} \\
        A = \fordis \alpha B A & \blam \alpha {\andcoerce{A}}
  \end{cases}
  \]
  \caption{Coercion generation considering Top-like types.}
  \label{fig:andcoercion}
\end{figure*}



%\paragraph{Atomic Types.} The new system introduces atomic types. Essentially a type
%is atomic if it is any type, which is not an intersection type.
%The notion of atomic
%type will be helpful

\paragraph{Well-formedness.}
We require that the two types of an intersection must be disjoint in their
context, and that the disjointness constraint in a universal type is well-formed.
Under the new rules, intersection types such as $\tyint \inter \tyint$ are no
longer well-formed because the two types are not disjoint.

\paragraph{Disjoint quantification.} A disjoint quantification is introduced by
the big lambda $\blamdis \alpha A e$ and eliminated by the usual type
application $\tapp e A$. The constraint is added to the context with this rule.
During a type application, the type system makes sure that the type argument
satisfies the disjointness constraint.

\paragraph{Metatheory.}

Since in this section we only restrict the type system in the previous section,
it is easy to see that type preservation and type-safety still holds.
Additionally, we can show that typing always produces a well-formed type by
proving the following results.

\begin{restatable}[Instantiation]{lemma}{instantiation}
  \label{lemma:instantiation}

  If $\jwf {\Gamma, \alpha \disjoint B} C$, $\jwf \Gamma A$, $\jdis \Gamma A B$
  then $\jwf \Gamma {\subst A \alpha C}$.
\end{restatable}

\begin{restatable}[Well-formed typing]{lemma}{wellformedtyping}
  \label{lemma:wellformed-typing}

  If $\jtype \Gamma e A$, then $\jwf \Gamma A$.
\end{restatable}

\begin{proof}
  By induction on the derivation that leads to $\jtype \Gamma e A$ and applying
  Lemma~\ref{lemma:instantiation} in the case of \reflabel{\labelttapp}.
\end{proof}

With our new definition of well-formed types, this result is nontrivial.
In general, disjointness judgements are not invariant with respect to
free-variable substitution. In other words, a careless substitution can violate
the disjoint constraint in the context. For example, in the context $\alpha
\disjoint \tyint$, $\alpha$ and $\tyint$ are disjoint:
\[ \jdis {\alpha \disjoint \tyint} \alpha \tyint \]
But after the substitution of $\tyint$ for $\alpha$ on the two types, the sentence
\[ \jdis {\alpha \disjoint \tyint} \tyint \tyint \]
is longer true since $\tyint$ is clearly not disjoint with itself.

\subsection{Subtyping}

% \george{Explain \reflabel{\labelsubforall} and distinction of kernel and all version.}

The subtyping rules need some adjustment.
Note that the $\bot$ type does not participate in subtyping since it holds no value.
An important
problem with the subtyping rules in Figure~\ref{fig:fi-type} is that the all three rules
dealing with intersection types
(\reflabel{\labelsubinterl} and \reflabel{\labelsubinterr} and \reflabel{\labelsubinter})
overlap. Unfortunatelly,
this means that different coercions may be given when checking the subtyping
between two types, depending on which derivation is chosen. This is ultimately the reason
for incoherence.
There are two important types of overlap:

\begin{enumerate}

\item The left decomposition rules for intersections (\reflabel{\labelsubinterl} and \reflabel{\labelsubinterr}) overlap with each other.

\item The left decomposition rules for intersections (\reflabel{\labelsubinterl} and \reflabel{\labelsubinterr})
overlap with the right decomposition rules for intersections (\reflabel{\labelsubinter}).

\end{enumerate}

\noindent Fortunately, disjoint intersections (which are enforced by well-formedness)
deal with problem 1): only one of the two left decomposition rules
can be chosen for a disjoint intersection type. Since the two types in the intersection
are disjoint it is impossible that both of the preconditions of the left decompositions are satisfied
at the same time. More formally, with disjoint intersections, we have the following theorem:

\begin{lemma}[Unique subtype contributor]
  \label{lemma:unique-subtype-contributor}

  If $A_1 \inter A_2 \subtype B$, where $A_1 \inter A_2$ and $B$ are well-formed types,
  then it is not possible that the following holds at the same time:
  \begin{enumerate}
    \item $A_1 \subtype B$
    \item $A_2 \subtype B$
  \end{enumerate}
\end{lemma}

Unfortunatelly, disjoint intersections alone are insufficient to deal with problem 2).
In order to deal with problem 2), we introduce a distinction between types, and atomic types.

\paragraph{Atomic types.} Atomic types are just those which are not intersection
types, and are asserted by the judgement \[ \jatomic A \]

In the left decomposition rules for intersections we introduce a requirement that
$A_3$ is atomic. The consequence of this requirement is that when $A_3$ is an intersection type, then
the only rule that can be applied is \reflabel{\labelsubinter}.
With the atomic constraint, one can guarantee that at any moment during the
derivation of a subtyping relation, at most one rule can be used.
Consequently, the coercion of a subtyping relation $A \subtype B$ is uniquely determined.
This fact is captured by the following lemma:

\begin{restatable}[Unique coercion]{lemma}{uniquecoercion}
  \label{lemma:unique-coercion}

  If $A \subtype B \yields {E_1}$ and $A \subtype B \yields {E_2}$, where $A$
  and $B$ are well-formed types, then $E_1 \equiv E_2$.
\end{restatable}

\paragraph{Expressiveness.}
Remarkably, our restrictions on subtyping do not sacrifice the expressiveness of
subtyping since we have the following two theorems:
\begin{theorem}
  If $A_1 \subtype A_3$, then $A_1 \inter A_2 \subtype A_3$.
\end{theorem}
\begin{theorem}
If $A_2 \subtype A_3$, then $A_1 \inter A_2 \subtype A_3$.
\end{theorem}

The interpretation of the theorem is that: even though the premise is made more
strict by the atomic condition, we can still derive the every subtyping relation
in the unrestricted system.
% \george{Explain why the proof shows this.}

\begin{comment}
Note that $A$ \emph{exclusive} or $B$ is true if and only if their truth value
differ. Next, we are going to investigate the minimal requirement (necessary and
sufficient conditions) such that the theorem holds.

If $A_1$ and $A_2$ in this setting are the same, for example,
$\tyint \inter \tyint \subtype \tyint$, obviously the theorem will
not hold since both the left $\tyint$ and the right $\tyint$ are a
subtype of $\tyint$.

We can try to rule out such possibilities by making the requirement of
well-formedness stronger. This suggests that the two types on the sides of
$\inter$ should not ``overlap''. In other words, they should be ``disjoint''. It
is easy to determine if two base types are disjoint. For example, $\tyint$
and $\tyint$ are not disjoint. Neither do $\tyint$ and $\code{Nat}$.
Also, types built with different constructors are disjoint. For example,
$\tyint$ and $\tyint \to \tyint$. For function types, disjointness
is harder to visualize. But bear in the mind that disjointness can defined by
the very requirement that the theorem holds.


This result is captured more formally by the following lemma:
\end{comment}

% \george{Note that $\bot$ does not participate in subtyping and why (because the
% empty set intersecting the empty set is still empty).}

% \george{What's the variance of the disjoint constraint? C.f. bounded
% polymorphism.}

% \george{Two points are being made here: 1) nondisjoint intersections, 2) atomic
% types. Show an offending example for each?}

\subsection{Coherence of the Elaboration}
Combining the previous results, we are able to show the central theorem:

\begin{restatable}[Unique elaboration]{theorem}{uniqueelaboration}
  \label{theorem:unique-elaboration}

  If $\jtype \Gamma e {A_1} \yields {E_1}$ and $\jtype \Gamma e {A_2} \yields
  {E_2}$, then $E_1 \equiv E_2$. (``$\equiv$'' means syntactical equality, up to
  $\alpha$-equality.)

\end{restatable}

\begin{proof}
  Note that the typing rules are already syntax-directed but the case of
  \reflabel{\labeltapp} (copied below) still needs special attention since we
  need to show that the generated coercion $E$ is unique.
  \begin{mathpar}
    \ruletapp
  \end{mathpar}
  Luckily, by Lemma~\ref{lemma:wellformed-typing}, we know that typing
  judgements give well-formed types, and thus $\jwf \Gamma {A_1}$ and $\jwf
  \Gamma {A_3}$. Therefore we are able to apply
  Lemma~\ref{lemma:unique-coercion} and conclude that $E$ is unique.

\end{proof}
